---
title: "Holtemme chemistry–transcriptome integration"
author: "Camilo Escobar"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

# Analytical scope

This report characterises the Holtemme chemical-exposure gradient and tests one
predeclared transcriptome-wide association: whether site-centroid expression is
linearly aligned with total predicted chronic mixture pressure,
`log10_STU_ToxC`. Chemical-specific and mode-of-action (MoA) profiles describe
exposure composition but are not treated as parallel transcriptomic predictors.

The response contains every gene with nonzero variance across the 30 RNA-seq
libraries. This universe was selected because the prior gene-universe
sensitivity analysis gave the same nominal conclusion for all tested universes
and retained the all-gene site geometry (minimum distance correlation 0.915).
All five nonzero site-centroid principal components preserve the six-site
geometry exactly. Inference enumerates all 720 permutations of the chemistry
predictor. Coexpression-network construction is intentionally reserved for
Steps 07 and 08.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
options(stringsAsFactors = FALSE)
set.seed(1207)

project_dir <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
source(file.path(project_dir, "support", "workflow_helpers.R"))
sync_rstudio_pandoc()

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(purrr)
  library(stringr)
  library(ggplot2)
  library(matrixStats)
  library(vegan)
})

results_dir <- make_results_dir(
  project_dir, 2, "Chemistry_Transcriptome_Integration"
)
tables_dir <- file.path(results_dir, "tables")
plots_dir <- file.path(results_dir, "plots")
logs_dir <- file.path(results_dir, "logs")
chem_tables_dir <- file.path(tables_dir, "chemistry")
ordination_tables_dir <- file.path(tables_dir, "constrained_ordination")
chem_plots_dir <- file.path(plots_dir, "chemistry")
ordination_plots_dir <- file.path(plots_dir, "constrained_ordination")
invisible(lapply(
  c(
    results_dir, tables_dir, plots_dir, logs_dir, chem_tables_dir,
    ordination_tables_dir, chem_plots_dir, ordination_plots_dir
  ),
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
))

vst_file <- file.path(
  project_dir, "Results_01_Global_Transcriptome_Structure",
  "matrices", "vst_matrix_germany.tsv"
)
metadata_file <- file.path(
  project_dir, "Results_01_Global_Transcriptome_Structure",
  "tables", "metadata_matched_germany.csv"
)
station_file <- file.path(project_dir, "metadata", "station_crosswalk.csv")
stu_file <- file.path(project_dir, "02_inputs", "STU_log10STU_per_site.tsv")
chemical_file <- file.path(project_dir, "02_inputs", "TU_per_site-chemical.tsv")
moa_file <- file.path(project_dir, "02_inputs", "TU_per_site-MoA.tsv")
chemistry_seed_file <- file.path(
  project_dir, "metadata", "holtemme_chemistry_site_metrics.csv"
)

required_inputs <- c(
  vst_file, metadata_file, station_file, stu_file, chemical_file, moa_file
)
if (!all(file.exists(required_inputs))) {
  stop(
    "Missing required Step 02 inputs:\n",
    paste(required_inputs[!file.exists(required_inputs)], collapse = "\n")
  )
}

site_order <- paste0("H", 1:7)
transcriptomic_sites <- paste0("H", 2:7)

write_csv_safe <- function(x, path) {
  readr::write_csv(x, path, na = "")
  invisible(path)
}

save_plot_bundle <- function(plot, stem, width = 8, height = 5.5) {
  ggplot2::ggsave(
    paste0(stem, ".png"), plot, width = width, height = height,
    dpi = 320, bg = "white"
  )
  ggplot2::ggsave(
    paste0(stem, ".pdf"), plot, width = width, height = height,
    bg = "white"
  )
  invisible(stem)
}

theme_publication <- function(base_size = 11) {
  ggplot2::theme_minimal(base_size = base_size) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(face = "bold"),
      axis.title = ggplot2::element_text(face = "bold")
    )
}
```

# 1. Chemistry harmonisation

```{r chemistry-harmonisation}
station_tbl <- readr::read_csv(station_file, show_col_types = FALSE)
station_lookup <- station_tbl %>%
  dplyr::transmute(
    legacy_site = as.character(.data$legacy_site_id),
    station_tag = as.character(.data$station_tag),
    station_order = as.integer(.data$station_order),
    site_name = as.character(.data$site_name)
  )

stu <- readr::read_tsv(stu_file, show_col_types = FALSE) %>%
  dplyr::transmute(
    site = as.character(.data$Site),
    STU_ToxC = as.numeric(.data$STU_ToxC),
    log10_STU_ToxC = as.numeric(.data$log10_STU_ToxC)
  )

chemical_long <- readr::read_tsv(
  chemical_file, show_col_types = FALSE
) %>%
  dplyr::transmute(
    site = as.character(.data$sites),
    chemical = as.character(.data$chemical_name),
    moa = as.character(.data$MoA_broad_01),
    iTU = as.numeric(.data$iTU)
  ) %>%
  dplyr::group_by(.data$site, .data$chemical, .data$moa) %>%
  dplyr::summarise(iTU = sum(.data$iTU, na.rm = TRUE), .groups = "drop")

moa_long <- readr::read_tsv(moa_file, show_col_types = FALSE) %>%
  dplyr::transmute(
    site = as.character(.data$sites),
    moa = stringr::str_squish(as.character(.data$MoA_broad_01)),
    total_TU = as.numeric(.data$total_TU)
  ) %>%
  dplyr::group_by(.data$site, .data$moa) %>%
  dplyr::summarise(total_TU = sum(.data$total_TU, na.rm = TRUE), .groups = "drop")

positive_itu <- chemical_long$iTU[chemical_long$iTU > 0]
itu_pseudocount <- min(positive_itu, na.rm = TRUE) / 2
chemical_wide <- tidyr::expand_grid(
  site = site_order,
  chemical = sort(unique(chemical_long$chemical))
) %>%
  dplyr::left_join(
    chemical_long %>%
      dplyr::group_by(.data$site, .data$chemical) %>%
      dplyr::summarise(iTU = sum(.data$iTU), .groups = "drop"),
    by = c("site", "chemical")
  ) %>%
  dplyr::mutate(
    iTU = dplyr::coalesce(.data$iTU, 0),
    log_iTU = log10(.data$iTU + itu_pseudocount)
  ) %>%
  dplyr::select(.data$site, .data$chemical, .data$log_iTU) %>%
  tidyr::pivot_wider(
    names_from = .data$chemical,
    values_from = .data$log_iTU
  ) %>%
  dplyr::arrange(match(.data$site, site_order))

chemical_pca <- stats::prcomp(
  chemical_wide %>% dplyr::select(-.data$site),
  center = TRUE,
  scale. = TRUE
)
chemical_scores <- tibble::as_tibble(
  chemical_pca$x[, seq_len(min(3L, ncol(chemical_pca$x))), drop = FALSE]
) %>%
  dplyr::mutate(site = chemical_wide$site, .before = 1)

moa_wide <- tidyr::expand_grid(
  site = site_order,
  moa = sort(unique(moa_long$moa))
) %>%
  dplyr::left_join(moa_long, by = c("site", "moa")) %>%
  dplyr::mutate(total_TU = dplyr::coalesce(.data$total_TU, 0)) %>%
  tidyr::pivot_wider(
    names_from = .data$moa,
    values_from = .data$total_TU,
    values_fill = 0
  ) %>%
  dplyr::arrange(match(.data$site, site_order))

positive_moa <- unlist(moa_wide %>% dplyr::select(-.data$site))
positive_moa <- positive_moa[positive_moa > 0]
moa_pseudocount <- min(positive_moa, na.rm = TRUE) / 2
moa_matrix <- log10(
  as.matrix(moa_wide %>% dplyr::select(-.data$site)) + moa_pseudocount
)
moa_pca <- stats::prcomp(moa_matrix, center = TRUE, scale. = TRUE)
moa_scores <- tibble::as_tibble(
  moa_pca$x[, seq_len(min(3L, ncol(moa_pca$x))), drop = FALSE]
) %>%
  dplyr::mutate(site = moa_wide$site, .before = 1)

# Deterministic publication exports for the current SI S1/S2 objects.
# These writes expose the existing objects without changing preprocessing,
# filtering, PCA definitions, or MoA categories. Original SI composite
# raster code is not available in this workflow and is therefore not
# fabricated here; canonical source data are exported instead.
publication_artifacts_dir <- file.path(results_dir, "publication_artifacts")
s1_publication_dir <- file.path(publication_artifacts_dir, "Figure_S1")
s2_publication_dir <- file.path(publication_artifacts_dir, "Figure_S2")
invisible(lapply(c(s1_publication_dir, s2_publication_dir), dir.create,
  recursive = TRUE, showWarnings = FALSE
))

chemical_pca_scores_full <- tibble::as_tibble(chemical_pca$x) %>%
  dplyr::mutate(site = chemical_wide$site, .before = 1)
chemical_pca_loadings <- tibble::as_tibble(chemical_pca$rotation,
  rownames = "chemical"
)
chemical_use_class_panel_source <- chemical_long %>%
  dplyr::group_by(.data$site, .data$moa) %>%
  dplyr::summarise(iTU = sum(.data$iTU, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(match(.data$site, site_order), .data$moa)
write_csv_safe(chemical_pca_scores_full,
  file.path(s1_publication_dir, "chemical_pca_scores.csv")
)
write_csv_safe(chemical_pca_loadings,
  file.path(s1_publication_dir, "chemical_pca_loadings.csv")
)
write_csv_safe(chemical_use_class_panel_source,
  file.path(s1_publication_dir, "chemical_use_class_panel_source.csv")
)

moa_pca_scores_full <- tibble::as_tibble(moa_pca$x) %>%
  dplyr::mutate(site = moa_wide$site, .before = 1)
moa_pca_loadings <- tibble::as_tibble(moa_pca$rotation,
  rownames = "moa"
)
moa_heatmap_source <- moa_long %>%
  dplyr::arrange(match(.data$site, site_order), .data$moa)

chemical_summary <- chemical_long %>%
  dplyr::group_by(.data$site) %>%
  dplyr::summarise(
    chemical_richness = sum(.data$iTU > 0),
    dominant_chemical = .data$chemical[which.max(.data$iTU)],
    dominant_chemical_TU = max(.data$iTU),
    .groups = "drop"
  )
moa_summary <- moa_long %>%
  dplyr::group_by(.data$site) %>%
  dplyr::summarise(
    MoA_richness = sum(.data$total_TU > 0),
    dominant_MoA = .data$moa[which.max(.data$total_TU)],
    dominant_MoA_TU = max(.data$total_TU),
    .groups = "drop"
  )

chemistry_metrics <- tibble::tibble(site = site_order) %>%
  dplyr::left_join(stu, by = "site") %>%
  dplyr::left_join(chemical_scores, by = "site") %>%
  dplyr::rename_with(~paste0("chemical_TU_", .x), dplyr::starts_with("PC")) %>%
  dplyr::left_join(moa_scores, by = "site") %>%
  dplyr::rename_with(~paste0("MoA_TU_", .x), dplyr::starts_with("PC")) %>%
  dplyr::left_join(chemical_summary, by = "site") %>%
  dplyr::left_join(moa_summary, by = "site") %>%
  dplyr::left_join(station_lookup, by = c("site" = "station_tag")) %>%
  dplyr::mutate(
    station_tag = .data$site,
    exclude_from_expression_analysis = !.data$site %in% transcriptomic_sites,
    station_order = dplyr::if_else(
      .data$site == "H1", NA_integer_, .data$station_order
    )
  ) %>%
  dplyr::arrange(match(.data$site, site_order))

moa_log10_sumTU_panel_source <- chemistry_metrics %>%
  dplyr::select(
    .data$site, .data$station_order, .data$STU_ToxC,
    .data$log10_STU_ToxC
  )
write_csv_safe(moa_pca_scores_full,
  file.path(s2_publication_dir, "moa_pca_scores.csv")
)
write_csv_safe(moa_pca_loadings,
  file.path(s2_publication_dir, "moa_pca_loadings.csv")
)
write_csv_safe(moa_heatmap_source,
  file.path(s2_publication_dir, "moa_heatmap_source.csv")
)
write_csv_safe(moa_log10_sumTU_panel_source,
  file.path(s2_publication_dir, "moa_log10_sumTU_panel_source.csv")
)

write_csv_safe(
  chemistry_metrics,
  file.path(chem_tables_dir, "holtemme_chemistry_site_metrics.csv")
)
write_csv_safe(
  chemistry_metrics,
  chemistry_seed_file
)
write_csv_safe(chemical_long, file.path(chem_tables_dir, "chemical_TU_long.csv"))
write_csv_safe(moa_long, file.path(chem_tables_dir, "MoA_TU_long.csv"))

chemistry_site_tbl <- chemistry_metrics %>%
  dplyr::filter(!.data$exclude_from_expression_analysis) %>%
  dplyr::select(
    .data$site, .data$legacy_site, .data$station_tag, .data$station_order,
    .data$site_name, .data$STU_ToxC, .data$log10_STU_ToxC,
    .data$chemical_richness, .data$MoA_richness,
    .data$dominant_chemical, .data$dominant_MoA
  )
write_csv_safe(
  chemistry_site_tbl,
  file.path(chem_tables_dir, "site_chemistry_for_transcriptome_integration.csv")
)

knitr::kable(
  chemistry_site_tbl,
  digits = 4,
  caption = "Chemistry matched to the six transcriptomic sites."
)
```

```{r chemistry-figures, fig.width=9, fig.height=5}
chemistry_gradient <- chemistry_metrics %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = factor(.data$site, levels = site_order),
      y = .data$log10_STU_ToxC,
      group = 1
    )
  ) +
  ggplot2::geom_line(color = "grey45") +
  ggplot2::geom_point(size = 3.5, color = "#2c7fb8") +
  ggplot2::labs(
    title = "Total predicted chronic mixture pressure",
    x = "Site",
    y = expression(log[10](Sigma*TU))
  ) +
  theme_publication()
save_plot_bundle(
  chemistry_gradient,
  file.path(chem_plots_dir, "chemistry_exposure_gradient"),
  width = 9,
  height = 5
)
chemistry_gradient
```

```{r moa-composition, fig.width=10, fig.height=6}
moa_composition <- moa_long %>%
  dplyr::group_by(.data$site) %>%
  dplyr::mutate(proportion = .data$total_TU / sum(.data$total_TU)) %>%
  dplyr::ungroup() %>%
  ggplot2::ggplot(
    ggplot2::aes(
      factor(.data$site, levels = site_order),
      .data$proportion,
      fill = .data$moa
    )
  ) +
  ggplot2::geom_col() +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::labs(
    title = "Mode-of-action composition of summed toxic units",
    x = "Site",
    y = "Proportion",
    fill = "Mode of action"
  ) +
  theme_publication()
save_plot_bundle(
  moa_composition,
  file.path(chem_plots_dir, "MoA_TU_composition_stacked"),
  width = 10,
  height = 6
)
moa_composition
```

# 2. Complete-gene site-centroid response

```{r expression-response}
metadata <- readr::read_csv(metadata_file, show_col_types = FALSE) %>%
  dplyr::mutate(
    sample_code = as.character(.data$sample_code),
    station_tag = as.character(.data$station_tag)
  )
vst <- readr::read_tsv(vst_file, show_col_types = FALSE, progress = FALSE)
sample_columns <- intersect(metadata$sample_code, setdiff(names(vst), "gene_id"))
if (length(sample_columns) != 30L) {
  stop("Step 02 requires exactly 30 expression libraries.")
}
metadata <- metadata %>% dplyr::slice(match(sample_columns, .data$sample_code))
expression <- as.matrix(vst[, sample_columns])
storage.mode(expression) <- "double"
rownames(expression) <- as.character(vst$gene_id)
gene_variance <- matrixStats::rowVars(expression)
keep_gene <- is.finite(gene_variance) & gene_variance > 0
expression <- expression[keep_gene, , drop = FALSE]

centroid_gene_by_site <- vapply(
  transcriptomic_sites,
  function(current_site) {
    matrixStats::rowMeans2(
      expression,
      cols = which(metadata$station_tag == current_site)
    )
  },
  numeric(nrow(expression))
)
rownames(centroid_gene_by_site) <- rownames(expression)
colnames(centroid_gene_by_site) <- transcriptomic_sites
site_response <- t(centroid_gene_by_site)
site_feature_variance <- matrixStats::colVars(site_response)
site_response <- site_response[
  , is.finite(site_feature_variance) & site_feature_variance > 0,
  drop = FALSE
]

response_genes <- tibble::tibble(
  gene_id = colnames(site_response),
  site_centroid_variance = matrixStats::colVars(site_response)
) %>%
  dplyr::arrange(dplyr::desc(.data$site_centroid_variance))
write_csv_safe(
  response_genes,
  file.path(ordination_tables_dir, "site_centroid_rda_response_genes.csv")
)

rda_universe_decision <- tibble::tibble(
  selected_universe = "all_nonzero_variance",
  selected_gene_n = nrow(response_genes),
  sensitivity_rule = paste(
    "Use all nonzero-variance genes because all tested universes gave the",
    "same nominal conclusion and preserved all-gene centroid geometry."
  ),
  all_tested_universes_same_nominal_conclusion = TRUE,
  minimum_geometry_correlation = 0.9149187594140149,
  sensitivity_exact_p = 0.2986111111111111
)
write_csv_safe(
  rda_universe_decision,
  file.path(ordination_tables_dir, "rda_gene_universe_decision.csv")
)

expression_qc <- tibble::tibble(
  raw_vst_genes = nrow(vst),
  nonzero_variance_genes = nrow(expression),
  variable_site_centroid_genes = ncol(site_response),
  expression_libraries = length(sample_columns),
  transcriptomic_sites = nrow(site_response)
)
write_csv_safe(expression_qc, file.path(tables_dir, "expression_filter_summary.csv"))
knitr::kable(expression_qc, caption = "Expression response retained for RDA.")
```

# 3. Exact six-site redundancy analysis

```{r exact-rda}
site_chemistry <- chemistry_site_tbl %>%
  dplyr::slice(match(transcriptomic_sites, .data$station_tag))
rownames(site_response) <- transcriptomic_sites

site_pca <- stats::prcomp(site_response, center = TRUE, scale. = FALSE)
n_site_pcs <- min(5L, ncol(site_pca$x))
site_pc_response <- site_pca$x[, seq_len(n_site_pcs), drop = FALSE]
site_pc_variance <- tibble::tibble(
  axis = paste0("tx_PC", seq_len(n_site_pcs)),
  variance_explained = site_pca$sdev[seq_len(n_site_pcs)]^2 /
    sum(site_pca$sdev^2)
) %>%
  dplyr::mutate(cumulative_variance = cumsum(.data$variance_explained))
write_csv_safe(
  site_pc_variance,
  file.path(ordination_tables_dir, "site_centroid_transcriptome_PC_variance.csv")
)

all_permutations <- function(x) {
  if (length(x) == 1L) return(matrix(x, nrow = 1L))
  do.call(rbind, lapply(seq_along(x), function(i) {
    cbind(x[[i]], all_permutations(x[-i]))
  }))
}

exact_rda_test <- function(response, predictor) {
  centered_response <- scale(response, center = TRUE, scale = FALSE)
  centered_predictor <- predictor - mean(predictor)
  predictor_ss <- sum(centered_predictor^2)
  total_ss <- sum(centered_response^2)
  model_ss <- sum(crossprod(centered_predictor, centered_response)^2) /
    predictor_ss
  residual_df <- nrow(response) - 2L
  observed_f <- model_ss / ((total_ss - model_ss) / residual_df)
  permutation_matrix <- unique(all_permutations(predictor))
  permutation_centered <- sweep(
    permutation_matrix, 1, rowMeans(permutation_matrix), "-"
  )
  permutation_model_ss <- rowSums(
    (permutation_centered %*% centered_response)^2
  ) / predictor_ss
  permutation_f <- permutation_model_ss /
    ((total_ss - permutation_model_ss) / residual_df)
  list(
    pseudo_f = observed_f,
    exact_p = mean(permutation_f >= observed_f - sqrt(.Machine$double.eps)),
    permutation_n = nrow(permutation_matrix)
  )
}

predictor <- site_chemistry$log10_STU_ToxC
rda_model <- vegan::rda(
  site_response ~ predictor,
  scale = FALSE
)
r2 <- vegan::RsquareAdj(rda_model)
exact_test <- exact_rda_test(site_pc_response, predictor)

rda_summary <- tibble::tibble(
  model = "log10_STU_ToxC",
  predictor = "log10_STU_ToxC",
  response = paste0(
    "site_centroid_all_nonzero_variance_VST_genes_n", ncol(site_response)
  ),
  response_type = "complete_site_centroid_vst_expression",
  constrained_r2 = unname(r2$r.squared),
  adjusted_r2 = unname(r2$adj.r.squared),
  overall_F = exact_test$pseudo_f,
  overall_p = exact_test$exact_p,
  permutation_method = "exact enumeration of all predictor permutations",
  permutation_n = exact_test$permutation_n,
  minimum_attainable_p = 1 / exact_test$permutation_n,
  n_sites = nrow(site_response),
  n_response_features = ncol(site_response)
)

site_scores <- as.data.frame(
  vegan::scores(rda_model, display = "sites", choices = 1:2, scaling = 2)
) %>%
  tibble::rownames_to_column("station_tag")
names(site_scores)[2:3] <- c("axis1", "axis2")
site_scores <- site_scores %>%
  dplyr::left_join(
    site_chemistry %>%
      dplyr::select(
        .data$station_tag, .data$site_name, .data$station_order,
        .data$STU_ToxC, .data$log10_STU_ToxC
      ),
    by = "station_tag"
  )

permutation_table <- rda_summary %>%
  dplyr::transmute(
    term = .data$predictor,
    test_scope = "overall",
    pseudo_F = .data$overall_F,
    exact_permutation_p = .data$overall_p,
    enumerated_permutation_n = .data$permutation_n,
    minimum_attainable_p = .data$minimum_attainable_p
  )

write_csv_safe(
  rda_summary,
  file.path(ordination_tables_dir, "chemistry_rda_model_comparison.csv")
)
write_csv_safe(
  site_scores,
  file.path(ordination_tables_dir, "chemistry_rda_site_scores.csv")
)
write_csv_safe(
  permutation_table,
  file.path(ordination_tables_dir, "chemistry_rda_permutation_tests.csv")
)
write_csv_safe(
  rda_summary,
  file.path(ordination_tables_dir, "site_centroid_rda_summary.csv")
)
write_csv_safe(
  site_scores,
  file.path(ordination_tables_dir, "site_centroid_rda_stu_scores.csv")
)

knitr::kable(
  rda_summary,
  digits = 4,
  caption = "Exact RDA of the complete site-centroid transcriptome."
)
```

```{r rda-figure, fig.width=8, fig.height=6}
rda_plot <- ggplot2::ggplot(
  site_scores,
  ggplot2::aes(.data$axis1, .data$axis2)
) +
  ggplot2::geom_path(ggplot2::aes(group = 1), color = "grey65") +
  ggplot2::geom_point(
    ggplot2::aes(fill = .data$log10_STU_ToxC),
    shape = 21,
    size = 5
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = .data$station_tag),
    nudge_y = 0.05,
    fontface = "bold"
  ) +
  ggplot2::scale_fill_gradient2(
    low = "#4575b4",
    mid = "#ffffbf",
    high = "#d73027",
    midpoint = median(site_scores$log10_STU_ToxC),
    name = expression(log[10](Sigma*TU))
  ) +
  ggplot2::labs(
    title = "Site-centroid transcriptome constrained by total mixture pressure",
    subtitle = paste0(
      "Adjusted R² = ", round(rda_summary$adjusted_r2, 3),
      "; exact p = ", round(rda_summary$overall_p, 3),
      "; 720 permutations"
    ),
    x = "Constrained axis",
    y = "First unconstrained axis"
  ) +
  theme_publication()
save_plot_bundle(
  rda_plot,
  file.path(ordination_plots_dir, "rda_site_centroid_vst_log10_STU_ToxC"),
  width = 8,
  height = 6
)
rda_plot
```

# Interpretation

The complete site-centroid transcriptome explained `r
round(100 * rda_summary$constrained_r2, 1)`% of the unadjusted multivariate
variation along the total-pressure gradient, but the adjusted proportion was
`r round(100 * rda_summary$adjusted_r2, 1)`% and the exact permutation test was
not significant (`p = ` `r round(rda_summary$overall_p, 3)`). Thus, these six
sites do not provide evidence for a global linear transcriptome–mixture-pressure
alignment. This result does not exclude local, nonlinear, compound-specific, or
mode-of-action-specific responses.

```{r final-qc}
quality_control <- tibble::tibble(
  check_id = c(
    "six_transcriptomic_sites",
    "five_libraries_per_site",
    "all_nonzero_variance_genes_used",
    "complete_pc_geometry",
    "all_720_permutations",
    "single_chemistry_predictor"
  ),
  passed = c(
    nrow(site_response) == 6L,
    all(table(metadata$station_tag) == 5L),
    ncol(site_response) == nrow(expression),
    n_site_pcs == 5L && tail(site_pc_variance$cumulative_variance, 1) > 0.999999,
    exact_test$permutation_n == 720L,
    identical(rda_summary$predictor, "log10_STU_ToxC")
  )
)
quality_control <- quality_control %>%
  dplyr::mutate(status = dplyr::if_else(.data$passed, "PASS", "FAIL"))
write_csv_safe(
  quality_control,
  file.path(tables_dir, "step02_quality_control_gates.csv")
)
if (any(!quality_control$passed)) {
  stop("One or more Step 02 quality-control gates failed.")
}
knitr::kable(quality_control, caption = "Step 02 quality-control gates.")
```

```{r session-info}
writeLines(
  capture.output(sessionInfo()),
  file.path(logs_dir, "session_info.txt")
)
sessionInfo()
```

