---
title: "Holtemme GOseq target-enrichment seed preparation"
author: "Camilo Escobar"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
project_dir <- normalizePath(Sys.getenv("HOLTEMME_PROJECT_DIR", unset = getwd()), winslash = "/")
handoff <- file.path(project_dir, "Results_04_EchoGO_Seed_Preparation", "echogo_handoff", "manifests", "goseq_echogo_handoff_manifest.csv")
if (file.exists(handoff)) manifest <- utils::read.csv(handoff, stringsAsFactors = FALSE) else manifest <- data.frame()
```

# Scope

Step 04 defines the deterministic target GOseq handoff for the six validated
DESeq2 contrasts. It preserves the tested gene universe, contrast-specific DEG
foregrounds, median transcript-length bias covariates, primary eggNOG GO
mapping, and retained sensitivity mappings. The handoff is consumed unchanged
by the annotation-context evidence layer in Step 05.

```{r handoff}
if (nrow(manifest)) knitr::kable(manifest, caption = "GOseq-to-EchoGO handoff manifest")
```

The output is a reproducible seed layer: each contrast retains its tested
universe, significant foreground, GO identifiers, ontology labels, and source
hashes. No context requests are made in this step.
