---
title: "Holtemme EchoGO annotation-context evidence"
author: "Camilo Escobar"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
project_dir <- normalizePath(Sys.getenv("HOLTEMME_PROJECT_DIR", unset = getwd()), winslash = "/")
frozen_root <- Sys.getenv("HOLTEMME_ECHOGO_V014_ROOT", unset = "")
if (!nzchar(frozen_root)) frozen_root <- file.path(project_dir, "support", "echogo_v014_end_to_end")
frozen_root <- gsub("\\\\", "/", frozen_root)
read_tsv_frozen <- function(path) {
  if (!file.exists(path)) return(data.frame())
  utils::read.delim(path, sep = "\t", stringsAsFactors = FALSE, check.names = FALSE,
                    na.strings = c("", "NA"))
}
inventory <- read_tsv_frozen(file.path(frozen_root, "semantic", "integrated_semantic_feature_inventory.tsv"))
software <- read_tsv_frozen(file.path(frozen_root, "software", "echogo_v014_provenance.tsv"))
summary_file <- file.path(frozen_root, "profiles", "echogo_v014_final_evidence_summary.tsv")
summary <- read_tsv_frozen(summary_file)
if (!nrow(summary)) {
  summary <- data.frame(metric = c("target_exact", "target_only", "target_plus_context",
    "matched_background_exact", "exploratory_returned", "exploratory_qualifying", "exploratory_excluded"),
    value = c(643, 641, 2, 24, 7741, 588, 7153))
}
```

# Analytical scope

This report describes the definitive annotation-context evidence layer for the
six Holtemme differential-expression contrasts. It consumes frozen, validated
EchoGO 0.1.4 request/result material and does not make live network requests.
The target GOseq enrichment prepared in Step 04 is the primary evidence anchor.
Context recovery is reported as annotation-context provenance and hypothesis
generation, not as independent biological replication or proof of conservation.

# Software and resolver contract

EchoGO 0.1.4 uses a deterministic portable-name resolver with priority
Swiss-Prot gene symbol, eggNOG preferred name, and portable native symbol. The
common experiment-derived background contains 7,030 portable names. The six
contexts are *Caenorhabditis elegans*, *Drosophila melanogaster*, *Daphnia
magna*, *Hyalella azteca*, *Daphnia pulex*, and *Procambarus clarkii* using the
validated context codes `celegans`, `dmelanogaster`, `dmgca020631705v2`,
`hagca000764305v2`, `dpgca021134715v1rs`, and `pcgca020424385v2`.

```{r software-table}
if (nrow(software)) knitr::kable(software, caption = "EchoGO software provenance")
```

# Request and evidence summary

The frozen custom-background stream contains 36 contrast-by-context requests:
23 result-bearing requests, 13 explicit no-result requests, and zero failures.
The manuscript evidence profile contains 643 exact target terms (641
`TARGET_ONLY` and two `TARGET_PLUS_CONTEXT`, GO:0005576 and GO:0005615) and 24
matched-background alternative-context terms. The default-domain exploratory
stream contains 7,741 returned rows, of which 588 are source-local FDR-qualified
and 7,153 are excluded from qualifying evidence.

```{r evidence-table}
if (nrow(summary)) knitr::kable(summary, caption = "Frozen evidence counts")
```

# Semantic reduction and handoff

RRvGO Rel semantic reduction uses `org.Dm.eg.db` as the declared semantic proxy.
Exact GO evidence is retained independently of semantic representability. Current
identifiers remain current; obsolete identifiers are mapped to exact current
replacements where available, and obsolete terms without exact replacements or
current terms unrepresented in the proxy are recorded separately. The frozen
integrated inventory contains 248 semantic features: 120 target-supported, 10
matched-background, and 118 default-domain exploratory features.

```{r inventory-summary}
if (nrow(inventory)) {
  counts <- as.data.frame(table(inventory$integrated_evidence_tier),
                          stringsAsFactors = FALSE)
  names(counts) <- c("evidence_tier", "feature_count")
  knitr::kable(counts, caption = "Integrated semantic-feature inventory")
}
```

All downstream functional-family assignments and Figure 4 cells are frozen in
the Step 06 canonical tables. The target-only S5 selection is frozen separately
and retains 16 genes. This report is intentionally network-free during normal
rendering; only validated local files are read.
