---
title: "Holtemme functional synthesis"
output:
  html_document:
    toc: true
    number_sections: true
params:
  project_dir: !r Sys.getenv("HOLTEMME_PROJECT_DIR", ".")
  figure_output_dir: !r Sys.getenv("HOLTEMME_FIGURE_OUTPUT_DIR", "")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
project_dir <- normalizePath(params$project_dir, mustWork = TRUE)
release_root <- if (dir.exists(file.path(project_dir, "DP05_functional_evidence"))) project_dir else dirname(project_dir)
candidate_sources <- c(
  file.path(project_dir, "DP05_functional_evidence", "provenance", "figure4", "figure4_family_cell_provenance.tsv"),
  file.path(release_root, "DP05_functional_evidence", "provenance", "figure4", "figure4_family_cell_provenance.tsv"),
  file.path(project_dir, "support", "figure4", "figure4_family_cell_provenance.tsv"),
  file.path(project_dir, "support", "echogo_v014_end_to_end", "figure4", "figure4_v014_integrated_family_cell_source.tsv")
)
figure_source <- candidate_sources[file.exists(candidate_sources)][1L]
if (is.na(figure_source)) stop("No frozen Figure 4 family-cell provenance table was found.")
figure_output_dir <- params$figure_output_dir
if (!nzchar(figure_output_dir)) {
  figure_output_dir <- if (dir.exists(file.path(project_dir, "DP05_functional_evidence", "publication_artifacts", "Figure_4"))) file.path(project_dir, "DP05_functional_evidence", "publication_artifacts", "Figure_4") else if (dir.exists(file.path(release_root, "DP05_functional_evidence", "publication_artifacts", "Figure_4"))) file.path(release_root, "DP05_functional_evidence", "publication_artifacts", "Figure_4") else file.path(project_dir, "Results_06_Functional_Synthesis", "final_paper_figures")
}
dir.create(figure_output_dir, recursive = TRUE, showWarnings = FALSE)
```

# Scope and frozen input

This report reconstructs the publication Figure 4 from the frozen v0.1.4
family-cell provenance table. The table is the scientific input; the plotting
code does not rerun enrichment, semantic reduction, resolver calls, or family
assignment. Target-supported, matched-background, and default-domain
exploratory evidence remain separate provenance tiers.

```{r figure4-input}
figure4_family_cells <- read.delim(figure_source, sep = "\t", check.names = FALSE, stringsAsFactors = FALSE)
figure4_family_cells |> dplyr::summarise(
  family_cells = dplyr::n(),
  displayed_cells = sum(toupper(display_in_figure) %in% c("TRUE", "T", "1")),
  target_features = sum(as.integer(n_target_supported_features)),
  matched_features = sum(as.integer(n_matched_background_hypothesis_features)),
  exploratory_features = sum(as.integer(n_default_domain_exploratory_only_features))
)
```

# Figure 4: functional-response landscape

The complete reviewed construction and save logic is kept in the repository
script `scripts/reconstruct_figure4_from_frozen_provenance.R`. The Rmd calls
that same code path, writes the canonical PDF/PNG/SVG bundle, and leaves the
publication figure embedded in this report.

```{r build-figure4, fig.width=13.5, fig.height=15.2, out.width="100%"}
figure4_script <- file.path(project_dir, "scripts", "reconstruct_figure4_from_frozen_provenance.R")
if (!file.exists(figure4_script)) stop("Figure 4 reconstruction script is missing: ", figure4_script)
status <- system2(
  file.path(R.home("bin"), "Rscript.exe"),
  args = c(figure4_script, figure_source, figure_output_dir)
)
if (!identical(status, 0L)) stop("Figure 4 reconstruction failed")
figure4_png <- file.path(figure_output_dir, "figure4_functional_response_landscape.png")
figure4_pdf <- file.path(figure_output_dir, "figure4_functional_response_landscape.pdf")
figure4_svg <- file.path(figure_output_dir, "figure4_functional_response_landscape.svg")
stopifnot(file.exists(figure4_png), file.exists(figure4_pdf), file.exists(figure4_svg))
knitr::include_graphics(figure4_png)
```

**Figure 4 legend.** Bubble area represents the number of distinct semantic
features in each reviewed family cell. Nested fills preserve the evidence
hierarchy: target-supported Gammarus evidence, matched-background
annotation-context hypotheses, and default-domain exploratory hypotheses.
Annotation-context evidence is hypothesis-generating and is not independent
biological replication or cross-species validation.

```{r figure4-output-check}
data.frame(
  source = basename(figure_source),
  pdf = basename(figure4_pdf),
  png = basename(figure4_png),
  svg = basename(figure4_svg),
  png_md5 = unname(tools::md5sum(figure4_png))
)
```

# Representative genes linked to functional-response families

The candidate-gene panel is the gene-level companion to the Biological Process
functional synthesis. It contains only final target-supported EchoGO v0.1.4
candidates with traceable native-ID to GO term to semantic-feature to
functional-family provenance. The panel is illustrative and is not an
independent validation set.

```{r candidate-gene-panel, results='asis', fig.width=13.5, fig.height=8.8, out.width='100%'}
candidate_file <- c(
  file.path(project_dir, "DP05_functional_evidence", "publication_artifacts", "Figure_S5", "candidate_gene_display_panel.tsv"),
  file.path(release_root, "DP05_functional_evidence", "publication_artifacts", "Figure_S5", "candidate_gene_display_panel.tsv"),
  file.path(project_dir, "support", "echogo_v014_end_to_end", "s5", "candidate_gene_display_panel.tsv")
)
candidate_file <- candidate_file[file.exists(candidate_file)][1L]
candidate_mat_file <- c(
  file.path(project_dir, "DP05_functional_evidence", "publication_artifacts", "Figure_S5", "candidate_gene_expression_site_mean_row_scaled_vst.tsv"),
  file.path(release_root, "DP05_functional_evidence", "publication_artifacts", "Figure_S5", "candidate_gene_expression_site_mean_row_scaled_vst.tsv")
)
candidate_mat_file <- candidate_mat_file[file.exists(candidate_mat_file)][1L]
candidate_heatmap_file <- c(
  file.path(project_dir, "DP05_functional_evidence", "publication_artifacts", "Figure_S5", "figure5_functional_family_linked_candidate_gene_heatmap.png"),
  file.path(release_root, "DP05_functional_evidence", "publication_artifacts", "Figure_S5", "figure5_functional_family_linked_candidate_gene_heatmap.png")
)
candidate_heatmap_file <- candidate_heatmap_file[file.exists(candidate_heatmap_file)][1L]
if (is.na(candidate_file) || is.na(candidate_mat_file) || is.na(candidate_heatmap_file)) {
  cat("The frozen 16-gene S5 source is not present in this checkout. The release candidate carries the machine-readable panel and heatmap.")
} else {
  candidate_panel <- read.delim(candidate_file, sep = "\t", check.names = FALSE, stringsAsFactors = FALSE)
  candidate_mat <- read.delim(candidate_mat_file, sep = "\t", check.names = FALSE, stringsAsFactors = FALSE)
  stopifnot(nrow(candidate_panel) == 16L, nrow(candidate_mat) == 16L)
  candidate_panel <- candidate_panel[order(candidate_panel$selection_rank_global), , drop = FALSE]
  candidate_mat <- candidate_mat[match(candidate_panel$heatmap_row_label, candidate_mat$heatmap_row_label), , drop = FALSE]
  candidate_families <- unique(candidate_panel$primary_family)
  cat("The final panel contains **16 displayed genes** across **", length(candidate_families),
      " Biological Process families**, using H2-H7 site means derived from **30 individual libraries**. ",
      "All displayed rows are target-supported evidence only.\n\n", sep = "")
  cat(sprintf("<p><img src=\"%s\" alt=\"Final 16-gene S5 heatmap\" style=\"display:block; margin:auto; width:100%%;\" /></p>", base64enc::dataURI(file = candidate_heatmap_file, mime = "image/png")))
  compact_candidate_table <- candidate_panel[, c("candidate_display_label", "primary_family", "family_contrasts"), drop = FALSE]
  names(compact_candidate_table) <- c("Gene", "Functional family", "Principal linked contrast or provenance context")
  print(knitr::kable(compact_candidate_table, caption = "Final 16-gene target-supported S5 panel"))
  cat("Colours are within-gene relative expression: red indicates relatively higher expression for a gene across sites and blue indicates relatively lower expression. Candidate selection is based on traceable target-supported functional evidence. The heatmap does not establish biomarkers, causal chemical drivers, pathway activation magnitude, or regulatory networks.")
}
```
