---
title: "Holtemme functional synthesis"
output:
  html_document:
    toc: true
    number_sections: true
params:
  project_dir: !r Sys.getenv("HOLTEMME_PROJECT_DIR", ".")
  figure_output_dir: !r Sys.getenv("HOLTEMME_FIGURE_OUTPUT_DIR", "")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
project_dir <- normalizePath(params$project_dir, mustWork = TRUE)
release_root <- if (dir.exists(file.path(project_dir, "DP05_functional_evidence"))) project_dir else dirname(project_dir)
candidate_sources <- c(
  file.path(project_dir, "DP05_functional_evidence", "provenance", "figure4", "figure4_family_cell_provenance.tsv"),
  file.path(release_root, "DP05_functional_evidence", "provenance", "figure4", "figure4_family_cell_provenance.tsv"),
  file.path(project_dir, "support", "figure4", "figure4_family_cell_provenance.tsv"),
  file.path(project_dir, "support", "echogo_v014_end_to_end", "figure4", "figure4_v014_integrated_family_cell_source.tsv")
)
figure_source <- candidate_sources[file.exists(candidate_sources)][1L]
if (is.na(figure_source)) stop("No frozen Figure 4 family-cell provenance table was found.")
figure_output_dir <- params$figure_output_dir
if (!nzchar(figure_output_dir)) {
  figure_output_dir <- if (dir.exists(file.path(project_dir, "DP05_functional_evidence", "publication_artifacts", "Figure_4"))) file.path(project_dir, "DP05_functional_evidence", "publication_artifacts", "Figure_4") else if (dir.exists(file.path(release_root, "DP05_functional_evidence", "publication_artifacts", "Figure_4"))) file.path(release_root, "DP05_functional_evidence", "publication_artifacts", "Figure_4") else file.path(project_dir, "Results_06_Functional_Synthesis", "final_paper_figures")
}
dir.create(figure_output_dir, recursive = TRUE, showWarnings = FALSE)
```

# Scope and frozen input

This report reconstructs the publication Figure 4 from the frozen v0.1.4
family-cell provenance table. The table is the scientific input; the plotting
code does not rerun enrichment, semantic reduction, resolver calls, or family
assignment. Target-supported, matched-background, and default-domain
exploratory evidence remain separate provenance tiers.

```{r figure4-input}
figure4_family_cells <- read.delim(figure_source, sep = "\t", check.names = FALSE, stringsAsFactors = FALSE)
figure4_family_cells |> dplyr::summarise(
  family_cells = dplyr::n(),
  displayed_cells = sum(toupper(display_in_figure) %in% c("TRUE", "T", "1")),
  target_features = sum(as.integer(n_target_supported_features)),
  matched_features = sum(as.integer(n_matched_background_hypothesis_features)),
  exploratory_features = sum(as.integer(n_default_domain_exploratory_only_features))
)
```

# Figure 4: functional-response landscape

The complete reviewed construction and save logic is kept in the repository
script `scripts/reconstruct_figure4_from_frozen_provenance.R`. The Rmd calls
that same code path, writes the canonical PDF/PNG/SVG bundle, and leaves the
publication figure embedded in this report.

```{r build-figure4, fig.width=13.5, fig.height=15.2, out.width="100%"}
figure4_script <- file.path(project_dir, "scripts", "reconstruct_figure4_from_frozen_provenance.R")
if (!file.exists(figure4_script)) stop("Figure 4 reconstruction script is missing: ", figure4_script)
status <- system2(
  file.path(R.home("bin"), "Rscript.exe"),
  args = c(figure4_script, figure_source, figure_output_dir)
)
if (!identical(status, 0L)) stop("Figure 4 reconstruction failed")
figure4_png <- file.path(figure_output_dir, "figure4_functional_response_landscape.png")
figure4_pdf <- file.path(figure_output_dir, "figure4_functional_response_landscape.pdf")
figure4_svg <- file.path(figure_output_dir, "figure4_functional_response_landscape.svg")
stopifnot(file.exists(figure4_png), file.exists(figure4_pdf), file.exists(figure4_svg))
knitr::include_graphics(figure4_png)
```

**Figure 4 legend.** Bubble area represents the number of distinct semantic
features in each reviewed family cell. Nested fills preserve the evidence
hierarchy: target-supported Gammarus evidence, matched-background
annotation-context hypotheses, and default-domain exploratory hypotheses.
Annotation-context evidence is hypothesis-generating and is not independent
biological replication or cross-species validation.

```{r figure4-output-check}
data.frame(
  source = figure_source,
  pdf = figure4_pdf,
  png = figure4_png,
  svg = figure4_svg,
  png_md5 = unname(tools::md5sum(figure4_png))
)
```
