---
title: "Holtemme functional synthesis"
author: "Camilo Escobar"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
project_dir <- normalizePath(Sys.getenv("HOLTEMME_PROJECT_DIR", unset = getwd()), winslash = "/")
root <- file.path(project_dir, "Results_06_Functional_Synthesis", "tables", "figure4_functional_landscape")
read_tsv <- function(n) { p <- file.path(root, n); if (file.exists(p)) utils::read.delim(p, sep = "\t", stringsAsFactors = FALSE, check.names = FALSE) else data.frame() }
inventory <- read_tsv("figure4_semantic_feature_inventory.tsv")
cells <- read_tsv("figure4_family_cell_provenance.tsv")
assignments <- read_tsv("functional_family_reviewed_assignments.tsv")
```

# Functional evidence model

Step 06 integrates target-supported GOseq evidence, matched-background
annotation-context hypotheses, and default-domain exploratory hypotheses. Exact
GO evidence is retained independently of semantic representability. RRvGO Rel
reduction uses `org.Dm.eg.db` as the declared semantic proxy; obsolete terms
without exact replacements and current terms unrepresented in the proxy are
recorded separately.

The final semantic inventory contains 248 integrated features: 120
target-supported, 10 matched-background, and 118 default-domain exploratory.
The reviewed family table is neutral and contains one row per integrated feature.

```{r tables}
if (nrow(assignments)) knitr::kable(utils::head(assignments, 20), caption = "Final reviewed family assignments (first 20 rows)")
if (nrow(cells)) knitr::kable(utils::head(cells, 20), caption = "Figure 4 family-cell provenance (first 20 rows)")
```

# Publication outputs

Figure 4 contains 101 displayed family cells with mutually exclusive
compositions of 43 target-only, 18 target-plus-exploratory, one matched-only,
four matched-plus-exploratory, and 35 exploratory-only cells. Evidence geometry
is nested: outer total, middle target plus matched-background, inner target.
There is no score or recurrence encoding. GO:0005509 in H7 versus H2 (MF) is
assigned to “Small-molecule and cofactor binding” as default-domain exploratory
evidence. The target-only S5 candidate-gene panel contains 16 genes.
