---
title: "Holtemme WGCNA network selection"
author: "Camilo Escobar"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

# Purpose and decision framework

This report tests whether the current WGCNA network is sensitive to retaining the
5,000, 10,000, or 15,000 most variable genes. It reconstructs each network from
the same 30 individual RNA-seq libraries while keeping the correlation method,
network type, module-detection settings, and planned site contrasts fixed. The
soft-thresholding power is re-evaluated separately for each gene set because the
connectivity distribution changes when additional genes enter the network.

The comparison is not designed to select the network with the highest
scale-free-topology fit alone. The final gene set should balance network
connectivity, module assignment, preservation of modules across nested gene
sets, site-stratified resampling stability, and computational reproducibility.
Annotation coverage is a secondary practical criterion. Site-contrast
significance, DEG overlap, enrichment yield, and chemistry are downstream
sensitivity results and cannot determine network parameters. Gene-set
preservation in this report is a
feature-selection sensitivity analysis and does not replace the site-stratified
resampling analysis required to evaluate module robustness.

# 0. Setup and configuration

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

options(stringsAsFactors = FALSE)
set.seed(2212)

project_dir <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
source(file.path(project_dir, "support", "workflow_helpers.R"))
sync_rstudio_pandoc()

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(purrr)
  library(stringr)
  library(ggplot2)
  library(matrixStats)
  library(WGCNA)
})
source(file.path(project_dir, "support", "wgcna_finalisation_helpers.R"))

WGCNA::allowWGCNAThreads(nThreads = 2)

results_dir <- make_results_dir(
  project_dir,
  7,
  "WGCNA_Network_Selection"
)
tables_dir <- file.path(results_dir, "tables")
plots_dir <- file.path(results_dir, "plots")
objects_dir <- file.path(results_dir, "objects")
logs_dir <- file.path(results_dir, "logs")

invisible(lapply(
  c(results_dir, tables_dir, plots_dir, objects_dir, logs_dir),
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
))

# Gene-set sensitivity design.
gene_set_sizes <- c(5000L, 10000L, 15000L)
reference_gene_set_size <- 5000L
candidate_powers <- c(1:10, seq(12, 20, by = 2))
soft_threshold_target <- 0.80
reference_power_for_comparison <- 12
power_sensitivity_candidates <- c(12L, 14L, 16L, 18L, 20L)
power_22_archive_sensitivity <- 22L
power_resample_replicates <- 10L
power_resample_libraries_per_site <- 4L
power_sensitivity_seed <- 2412L
very_low_connectivity_threshold <- 1
adequate_median_connectivity_threshold <- 10
maximum_scale_free_fit_deficit <- 0.03
candidate_module_moderate_median_jaccard <- 0.50
candidate_module_moderate_minimum_jaccard <- 0.25
candidate_module_moderate_eigengene_agreement <- 0.80
reuse_power_sensitivity_cache <- TRUE

# Network settings evaluated for the final coexpression analysis.
network_type <- "signed"
correlation_type <- "bicor"
max_p_outliers <- 0.10
minimum_module_size <- 50L
merge_cut_height <- 0.25
deep_split <- 2
pam_respects_dendro <- FALSE

# A finite block size reduces peak memory use for the 10,000- and 15,000-gene
# networks. Set this to max(gene_set_sizes) only when sufficient RAM is
# available and a single-block comparison is required.
max_block_size <- 6000L
reuse_network_cache <- TRUE

site_order <- c("H2", "H3", "H4", "H5", "H6", "H7")

planned_contrasts <- tibble::tribble(
  ~contrast, ~contrast_panel, ~numerator_site, ~denominator_site, ~contrast_role,
  "HOL02_vs_HOL01", "First WWTP transition\nH3 vs H2", "HOL02", "HOL01",
  "first local WWTP transition",
  "HOL05_vs_HOL04", "Second WWTP transition\nH6 vs H5", "HOL05", "HOL04",
  "second local WWTP transition",
  "HOL06_vs_HOL01", "Furthest downstream site\nH7 vs H2", "HOL06", "HOL01",
  "downstream integrative comparison"
)

large_effect_abs_hedges_g <- 1.00
nominal_p_threshold <- 0.05
fdr_threshold <- 0.05

write_csv_safe <- function(x, path) {
  readr::write_csv(x, path, na = "")
  invisible(path)
}

save_plot_bundle <- function(plot, stem, width = 9, height = 6) {
  ggplot2::ggsave(
    paste0(stem, ".png"),
    plot,
    width = width,
    height = height,
    dpi = 320,
    bg = "white"
  )
  ggplot2::ggsave(
    paste0(stem, ".pdf"),
    plot,
    width = width,
    height = height,
    bg = "white"
  )
  if (requireNamespace("svglite", quietly = TRUE)) {
    ggplot2::ggsave(
      paste0(stem, ".svg"),
      plot,
      width = width,
      height = height,
      bg = "white"
    )
  }
  invisible(stem)
}

theme_publication <- function(base_size = 11) {
  ggplot2::theme_minimal(base_size = base_size) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(face = "bold"),
      axis.title = ggplot2::element_text(face = "bold"),
      strip.text = ggplot2::element_text(face = "bold"),
      legend.position = "right"
    )
}

safe_chr <- function(x) {
  x <- stringr::str_squish(as.character(x))
  x[!nzchar(x)] <- NA_character_
  x
}

safe_divide <- function(numerator, denominator) {
  dplyr::if_else(
    is.finite(denominator) & denominator > 0,
    numerator / denominator,
    NA_real_
  )
}

hedges_g_two_group <- function(x_numerator, x_denominator) {
  x_numerator <- as.numeric(x_numerator)
  x_denominator <- as.numeric(x_denominator)
  x_numerator <- x_numerator[is.finite(x_numerator)]
  x_denominator <- x_denominator[is.finite(x_denominator)]

  n_num <- length(x_numerator)
  n_den <- length(x_denominator)
  if (n_num < 2L || n_den < 2L) return(NA_real_)

  pooled_df <- n_num + n_den - 2L
  pooled_var <- (
    (n_num - 1L) * stats::var(x_numerator) +
      (n_den - 1L) * stats::var(x_denominator)
  ) / pooled_df

  if (!is.finite(pooled_var) || pooled_var <= 0) return(NA_real_)

  cohen_d <- (
    mean(x_numerator) - mean(x_denominator)
  ) / sqrt(pooled_var)

  correction_j <- 1 - 3 / (4 * (n_num + n_den) - 9)
  correction_j * cohen_d
}

select_soft_power <- function(soft_tbl, target = 0.80) {
  eligible <- soft_tbl %>%
    dplyr::filter(
      is.finite(.data$scale_free_fit),
      .data$scale_free_fit >= target
    )

  if (nrow(eligible) > 0L) {
    selected <- eligible %>%
      dplyr::slice_min(
        order_by = .data$power,
        n = 1,
        with_ties = FALSE
      )
    rule <- paste0("lowest power reaching R2 >= ", target)
  } else {
    selected <- soft_tbl %>%
      dplyr::filter(is.finite(.data$scale_free_fit)) %>%
      dplyr::arrange(
        dplyr::desc(.data$scale_free_fit),
        .data$power
      ) %>%
      dplyr::slice_head(n = 1)
    rule <- "highest available scale-free fit; target not reached"
  }

  if (nrow(selected) == 0L) {
    stop("No finite soft-threshold diagnostics were available.")
  }

  list(
    power = as.numeric(selected$power[[1]]),
    rule = rule,
    row = selected
  )
}

fit_one_module_contrast <- function(
  values,
  site_ids,
  numerator_site,
  denominator_site
) {
  contrast_df <- tibble::tibble(
    value = as.numeric(values),
    site_id = as.character(site_ids)
  ) %>%
    dplyr::filter(
      !is.na(.data$site_id),
      is.finite(.data$value)
    ) %>%
    dplyr::mutate(
      site_id = factor(
        .data$site_id,
        levels = unique(as.character(site_ids))
      )
    )

  numerator_values <- contrast_df$value[
    contrast_df$site_id == numerator_site
  ]
  denominator_values <- contrast_df$value[
    contrast_df$site_id == denominator_site
  ]

  if (
    length(numerator_values) < 2L ||
      length(denominator_values) < 2L
  ) {
    return(tibble::tibble(
      n_numerator = length(numerator_values),
      n_denominator = length(denominator_values),
      numerator_mean = mean(numerator_values, na.rm = TRUE),
      denominator_mean = mean(denominator_values, na.rm = TRUE),
      eigengene_difference = NA_real_,
      conf_low = NA_real_,
      conf_high = NA_real_,
      hedges_g = NA_real_,
      p_value = NA_real_
    ))
  }

  fit <- stats::lm(value ~ 0 + site_id, data = contrast_df)
  beta_hat <- stats::coef(fit)
  beta_vcov <- stats::vcov(fit)
  residual_df <- stats::df.residual(fit)
  numerator_coef <- paste0("site_id", numerator_site)
  denominator_coef <- paste0("site_id", denominator_site)

  if (!all(c(numerator_coef, denominator_coef) %in% names(beta_hat))) {
    stop("A planned contrast site is absent from the common six-site model.")
  }

  contrast_vector <- stats::setNames(
    rep(0, length(beta_hat)),
    names(beta_hat)
  )
  contrast_vector[[numerator_coef]] <- 1
  contrast_vector[[denominator_coef]] <- -1
  estimate <- sum(contrast_vector * beta_hat)
  standard_error <- sqrt(as.numeric(
    t(contrast_vector) %*% beta_vcov %*% contrast_vector
  ))
  t_statistic <- estimate / standard_error
  p_value <- 2 * stats::pt(-abs(t_statistic), df = residual_df)
  critical_t <- stats::qt(0.975, df = residual_df)

  tibble::tibble(
    n_numerator = length(numerator_values),
    n_denominator = length(denominator_values),
    numerator_mean = mean(numerator_values),
    denominator_mean = mean(denominator_values),
    eigengene_difference = estimate,
    conf_low = estimate - critical_t * standard_error,
    conf_high = estimate + critical_t * standard_error,
    hedges_g = hedges_g_two_group(
      numerator_values,
      denominator_values
    ),
    p_value = p_value
  )
}

fit_planned_module_contrasts <- function(
  eigengene_tbl,
  metadata_tbl,
  network_label,
  network_size,
  module_assignment,
  contrast_tbl
) {
  me_columns <- grep("^ME", names(eigengene_tbl), value = TRUE)
  me_columns <- setdiff(me_columns, "MEgrey")

  module_sizes <- module_assignment %>%
    dplyr::filter(.data$module != "grey") %>%
    dplyr::count(
      .data$module,
      .data$module_label,
      name = "module_size"
    )

  result <- purrr::map_dfr(
    me_columns,
    function(current_module_label) {
      purrr::pmap_dfr(
        contrast_tbl,
        function(
          contrast,
          contrast_panel,
          numerator_site,
          denominator_site,
          contrast_role
        ) {
          test_row <- fit_one_module_contrast(
            values = eigengene_tbl[[current_module_label]],
            site_ids = metadata_tbl$site_id,
            numerator_site = numerator_site,
            denominator_site = denominator_site
          )

          test_row %>%
            dplyr::mutate(
              network_label = network_label,
              network_size = network_size,
              module_label = current_module_label,
              module = stringr::str_remove(
                current_module_label,
                "^ME"
              ),
              contrast = contrast,
              contrast_panel = contrast_panel,
              numerator_site = numerator_site,
              denominator_site = denominator_site,
              contrast_role = contrast_role,
              .before = 1
            )
        }
      )
    }
  ) %>%
    dplyr::left_join(
      module_sizes,
      by = c("module", "module_label")
    ) %>%
    dplyr::mutate(
      q_global_network = stats::p.adjust(
        .data$p_value,
        method = "BH"
      )
    ) %>%
    dplyr::group_by(.data$contrast) %>%
    dplyr::mutate(
      q_within_contrast = stats::p.adjust(
        .data$p_value,
        method = "BH"
      )
    ) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      fdr_supported = .data$q_global_network < fdr_threshold,
      large_nominal_effect =
        abs(.data$hedges_g) >= large_effect_abs_hedges_g &
        .data$p_value < nominal_p_threshold
    )

  result
}

run_network_for_gene_set <- function(
  requested_gene_n,
  ranked_gene_ids,
  expression_matrix,
  metadata_aligned
) {
  network_label <- paste0("G", requested_gene_n)
  message("Starting ", network_label, " network")

  selected_gene_ids <- utils::head(
    ranked_gene_ids,
    min(requested_gene_n, length(ranked_gene_ids))
  )

  dat_expr <- t(
    expression_matrix[selected_gene_ids, , drop = FALSE]
  )
  rownames(dat_expr) <- metadata_aligned$sample_code

  gsg <- WGCNA::goodSamplesGenes(dat_expr, verbose = 0)
  if (!gsg$allOK) {
    dat_expr <- dat_expr[
      gsg$goodSamples,
      gsg$goodGenes,
      drop = FALSE
    ]
  }

  metadata_network <- metadata_aligned %>%
    dplyr::filter(.data$sample_code %in% rownames(dat_expr)) %>%
    dplyr::slice(match(rownames(dat_expr), .data$sample_code))

  sft <- WGCNA::pickSoftThreshold(
    dat_expr,
    powerVector = candidate_powers,
    networkType = network_type,
    corFnc = correlation_type,
    corOptions = list(maxPOutliers = max_p_outliers),
    verbose = 0
  )

  soft_tbl <- tibble::as_tibble(sft$fitIndices) %>%
    dplyr::transmute(
      network_label = network_label,
      requested_gene_n = requested_gene_n,
      retained_gene_n = ncol(dat_expr),
      power = .data$Power,
      scale_free_fit = .data$SFT.R.sq,
      slope = .data$slope,
      mean_connectivity = .data$mean.k.,
      median_connectivity = .data$median.k.,
      max_connectivity = .data$max.k.
    )

  power_decision <- select_soft_power(
    soft_tbl,
    target = soft_threshold_target
  )
  selected_power <- power_decision$power

  net <- WGCNA::blockwiseModules(
    dat_expr,
    power = selected_power,
    networkType = network_type,
    TOMType = network_type,
    corType = correlation_type,
    maxPOutliers = max_p_outliers,
    minModuleSize = minimum_module_size,
    mergeCutHeight = merge_cut_height,
    deepSplit = deep_split,
    numericLabels = FALSE,
    pamRespectsDendro = pam_respects_dendro,
    maxBlockSize = min(max_block_size, ncol(dat_expr)),
    saveTOMs = FALSE,
    verbose = 0
  )

  raw_module_colors <- as.character(net$colors)
  names(raw_module_colors) <- colnames(dat_expr)
  corrected_modules <- wgcna_correct_undersized_modules(
    raw_module_colors,
    minimum_module_size
  )
  module_colors <- corrected_modules$colors
  oriented_modules <- wgcna_orient_eigengenes(
    dat_expr,
    module_colors,
    max_p_outliers = max_p_outliers
  )
  module_eigengenes <- oriented_modules$eigengenes

  module_assignment <- tibble::tibble(
    network_label = network_label,
    network_size = requested_gene_n,
    gene_id = names(module_colors),
    module = unname(module_colors),
    module_label = paste0("ME", unname(module_colors))
  )

  kme_tbl <- tibble::as_tibble(
    WGCNA::signedKME(
      dat_expr,
      module_eigengenes,
      outputColumnName = "kME",
      corFnc = correlation_type,
      corOptions = paste0(
        "maxPOutliers = ",
        max_p_outliers
      )
    ),
    rownames = "gene_id"
  )

  module_assignment$module_kME_signed <- NA_real_
  for (current_module in unique(module_assignment$module)) {
    kme_column <- paste0("kME", current_module)
    current_rows <- module_assignment$module == current_module
    if (kme_column %in% names(kme_tbl)) {
      module_assignment$module_kME_signed[current_rows] <-
        kme_tbl[[kme_column]][
          match(
            module_assignment$gene_id[current_rows],
            kme_tbl$gene_id
          )
        ]
    }
  }

  module_assignment <- module_assignment %>%
    dplyr::mutate(
      module_kME_abs = abs(.data$module_kME_signed)
    )

  module_summary <- module_assignment %>%
    dplyr::count(
      .data$network_label,
      .data$network_size,
      .data$module,
      .data$module_label,
      name = "n_genes"
    ) %>%
    dplyr::mutate(
      module_type = dplyr::if_else(
        .data$module == "grey",
        "unassigned",
        "assigned"
      )
    ) %>%
    dplyr::arrange(
      .data$module_type,
      dplyr::desc(.data$n_genes)
    )

  assigned_module_sizes_for_assertion <- module_summary$n_genes[
    module_summary$module != "grey"
  ]
  if (length(assigned_module_sizes_for_assertion) > 0L &&
      any(assigned_module_sizes_for_assertion < minimum_module_size)) {
    stop(network_label, " retained an assigned module below minModuleSize.")
  }
  module_labels_from_colors <- paste0("ME", sort(unique(module_colors)))
  module_labels_from_mes <- names(module_eigengenes)
  if (!setequal(module_labels_from_colors, module_labels_from_mes)) {
    stop(network_label, " module-colour and eigengene labels do not match.")
  }

  eigengene_tbl <- tibble::as_tibble(
    module_eigengenes,
    rownames = "sample_code"
  ) %>%
    dplyr::left_join(
      metadata_network %>%
        dplyr::select(
          .data$sample_code,
          .data$site_id,
          .data$station_tag,
          dplyr::any_of(c(
            "display_label",
            "site_name",
            "station_order"
          ))
        ),
      by = "sample_code"
    )

  contrast_tbl <- fit_planned_module_contrasts(
    eigengene_tbl = eigengene_tbl,
    metadata_tbl = metadata_network,
    network_label = network_label,
    network_size = requested_gene_n,
    module_assignment = module_assignment,
    contrast_tbl = planned_contrasts
  )

  selected_soft_row <- power_decision$row

  assigned_sizes <- module_summary$n_genes[
    module_summary$module_type == "assigned"
  ]
  grey_n <- sum(
    module_summary$n_genes[module_summary$module == "grey"],
    na.rm = TRUE
  )
  assigned_n <- ncol(dat_expr) - grey_n

  network_summary <- tibble::tibble(
    network_label = network_label,
    requested_gene_n = requested_gene_n,
    retained_gene_n = ncol(dat_expr),
    retained_sample_n = nrow(dat_expr),
    selected_power = selected_power,
    selected_power_rule = power_decision$rule,
    selected_scale_free_fit = selected_soft_row$scale_free_fit[[1]],
    selected_slope = selected_soft_row$slope[[1]],
    selected_mean_connectivity =
      selected_soft_row$mean_connectivity[[1]],
    selected_median_connectivity =
      selected_soft_row$median_connectivity[[1]],
    target_reached =
      selected_soft_row$scale_free_fit[[1]] >= soft_threshold_target,
    reference_power_scale_free_fit = soft_tbl$scale_free_fit[
      match(reference_power_for_comparison, soft_tbl$power)
    ],
    reference_power_mean_connectivity = soft_tbl$mean_connectivity[
      match(reference_power_for_comparison, soft_tbl$power)
    ],
    n_assigned_modules = sum(
      module_summary$module_type == "assigned"
    ),
    n_assigned_genes = assigned_n,
    assigned_gene_fraction = assigned_n / ncol(dat_expr),
    n_grey_genes = grey_n,
    grey_gene_fraction = grey_n / ncol(dat_expr),
    median_assigned_module_size = if (
      length(assigned_sizes) > 0L
    ) stats::median(assigned_sizes) else NA_real_,
    minimum_assigned_module_size = if (
      length(assigned_sizes) > 0L
    ) min(assigned_sizes) else NA_real_,
    maximum_assigned_module_size = if (
      length(assigned_sizes) > 0L
    ) max(assigned_sizes) else NA_real_,
    n_blocks = length(net$blockGenes),
    max_block_size_setting = min(max_block_size, ncol(dat_expr)),
    n_fdr_supported_contrasts = sum(
      contrast_tbl$fdr_supported,
      na.rm = TRUE
    ),
    n_large_nominal_contrasts = sum(
      contrast_tbl$large_nominal_effect,
      na.rm = TRUE
    )
  )

  write_csv_safe(
    soft_tbl,
    file.path(
      tables_dir,
      paste0(network_label, "_soft_threshold_diagnostics.csv")
    )
  )
  write_csv_safe(
    module_summary,
    file.path(
      tables_dir,
      paste0(network_label, "_module_summary.csv")
    )
  )
  write_csv_safe(
    oriented_modules$orientation,
    file.path(tables_dir, paste0(network_label, "_eigengene_orientation.csv"))
  )
  write_csv_safe(
    tibble::tibble(
      raw_module = names(corrected_modules$raw_sizes),
      raw_module_size = as.integer(corrected_modules$raw_sizes)
    ) %>%
      dplyr::mutate(
        below_declared_minimum =
          .data$raw_module != "grey" &
          .data$raw_module_size < minimum_module_size,
        final_action = dplyr::if_else(
          .data$below_declared_minimum,
          "reclassified_to_exact_grey_before_testing",
          "retained"
        )
      ),
    file.path(tables_dir, paste0(network_label, "_module_identity_audit.csv"))
  )
  write_csv_safe(
    module_assignment,
    file.path(
      tables_dir,
      paste0(network_label, "_gene_module_membership.csv")
    )
  )
  write_csv_safe(
    eigengene_tbl,
    file.path(
      tables_dir,
      paste0(network_label, "_module_eigengenes_by_sample.csv")
    )
  )
  write_csv_safe(
    contrast_tbl,
    file.path(
      tables_dir,
      paste0(network_label, "_planned_module_contrasts.csv")
    )
  )

  rm(net, dat_expr, kme_tbl)
  invisible(gc())

  list(
    network_label = network_label,
    network_size = requested_gene_n,
    selected_gene_ids = selected_gene_ids,
    soft = soft_tbl,
    summary = network_summary,
    modules = module_summary,
    assignment = module_assignment,
    eigengenes = eigengene_tbl,
    orientation = oriented_modules$orientation,
    raw_undersized_modules = corrected_modules$undersized_modules,
    contrasts = contrast_tbl
  )
}

load_or_run_network_for_gene_set <- function(
  requested_gene_n,
  ranked_gene_ids,
  expression_matrix,
  metadata_aligned,
  input_signature
) {
  network_label <- paste0("G", requested_gene_n)
  selected_ids <- utils::head(
    ranked_gene_ids,
    min(requested_gene_n, length(ranked_gene_ids))
  )
  cache_signature <- paste(
    input_signature,
    paste0("network_label=", network_label),
    paste0(
      "selected_gene_hash=",
      digest::digest(selected_ids, algo = "xxhash64")
    ),
    paste0("WGCNA=", as.character(utils::packageVersion("WGCNA"))),
    paste0("powers=", paste(candidate_powers, collapse = ";")),
    paste0("target=", soft_threshold_target),
    paste0("network_type=", network_type),
    paste0("correlation_type=", correlation_type),
    paste0("max_p_outliers=", max_p_outliers),
    paste0("minimum_module_size=", minimum_module_size),
    paste0("merge_cut_height=", merge_cut_height),
    paste0("deep_split=", deep_split),
    paste0("pam_respects_dendro=", pam_respects_dendro),
    paste0("max_block_size=", max_block_size),
    "finalisation_validation=v2",
    sep = "||"
  )
  cache_file <- file.path(
    objects_dir,
    paste0(network_label, "_network_result_cache.rds")
  )

  if (reuse_network_cache && file.exists(cache_file)) {
    cached <- readRDS(cache_file)
    if (
      is.list(cached) &&
        identical(cached$cache_signature, cache_signature) &&
        is.list(cached$result)
    ) {
      message("Reusing validated cache for ", network_label)
      result <- cached$result
      result$contrasts <- fit_planned_module_contrasts(
        eigengene_tbl = result$eigengenes,
        metadata_tbl = metadata_aligned,
        network_label = result$network_label,
        network_size = result$network_size,
        module_assignment = result$assignment,
        contrast_tbl = planned_contrasts
      )
      result$summary <- result$summary %>%
        dplyr::mutate(
          n_fdr_supported_contrasts = sum(
            result$contrasts$fdr_supported,
            na.rm = TRUE
          ),
          n_large_nominal_contrasts = sum(
            result$contrasts$large_nominal_effect,
            na.rm = TRUE
          )
        )
      write_csv_safe(
        result$contrasts,
        file.path(
          tables_dir,
          paste0(network_label, "_planned_module_contrasts.csv")
        )
      )
      cached$result <- result
      saveRDS(cached, cache_file, compress = FALSE)
      return(result)
    }
  }

  result <- run_network_for_gene_set(
    requested_gene_n = requested_gene_n,
    ranked_gene_ids = ranked_gene_ids,
    expression_matrix = expression_matrix,
    metadata_aligned = metadata_aligned
  )
  saveRDS(
    list(cache_signature = cache_signature, result = result),
    cache_file,
    compress = FALSE
  )
  result
}

compare_network_pair <- function(reference_result, test_result) {
  reference_label <- reference_result$network_label
  test_label <- test_result$network_label

  reference_assignment <- reference_result$assignment %>%
    dplyr::select(
      gene_id,
      reference_module = module,
      reference_module_label = module_label
    )
  test_assignment <- test_result$assignment %>%
    dplyr::select(
      gene_id,
      test_module = module,
      test_module_label = module_label
    )

  shared_assignment <- reference_assignment %>%
    dplyr::inner_join(test_assignment, by = "gene_id")

  reference_modules <- shared_assignment %>%
    dplyr::filter(.data$reference_module != "grey") %>%
    dplyr::distinct(
      .data$reference_module,
      .data$reference_module_label
    )

  test_assigned_modules <- shared_assignment %>%
    dplyr::filter(.data$test_module != "grey") %>%
    dplyr::distinct(
      .data$test_module,
      .data$test_module_label
    )

  reference_eigengenes <- reference_result$eigengenes
  test_eigengenes <- test_result$eigengenes

  purrr::pmap_dfr(
    reference_modules,
    function(reference_module, reference_module_label) {
      reference_gene_set <- shared_assignment$gene_id[
        shared_assignment$reference_module == reference_module
      ]
      reference_to_grey_n <- sum(
        shared_assignment$reference_module == reference_module &
          shared_assignment$test_module == "grey"
      )

      candidate_matches <- purrr::pmap_dfr(
        test_assigned_modules,
        function(test_module, test_module_label) {
          test_gene_set <- shared_assignment$gene_id[
            shared_assignment$test_module == test_module
          ]
          intersection_n <- length(intersect(
            reference_gene_set,
            test_gene_set
          ))
          union_n <- length(union(
            reference_gene_set,
            test_gene_set
          ))

          tibble::tibble(
            test_module = test_module,
            test_module_label = test_module_label,
            intersection_n = intersection_n,
            union_n = union_n,
            jaccard = if (
              union_n > 0L
            ) intersection_n / union_n else NA_real_,
            reference_recovery_fraction = if (
              length(reference_gene_set) > 0L
            ) intersection_n / length(reference_gene_set) else NA_real_,
            test_purity_fraction = if (
              length(test_gene_set) > 0L
            ) intersection_n / length(test_gene_set) else NA_real_
          )
        }
      ) %>%
        dplyr::arrange(
          dplyr::desc(.data$jaccard),
          dplyr::desc(.data$intersection_n)
        )

      best_match <- candidate_matches %>%
        dplyr::slice_head(n = 1)

      if (nrow(best_match) == 0L) {
        return(tibble::tibble(
          reference_network = reference_label,
          test_network = test_label,
          reference_size = reference_result$network_size,
          test_size = test_result$network_size,
          reference_module = reference_module,
          reference_module_label = reference_module_label,
          test_module = NA_character_,
          test_module_label = NA_character_,
          reference_module_shared_n = length(reference_gene_set),
          reference_to_test_grey_n = reference_to_grey_n,
          reference_to_test_grey_fraction = safe_divide(
            reference_to_grey_n,
            length(reference_gene_set)
          ),
          intersection_n = 0L,
          union_n = NA_integer_,
          jaccard = NA_real_,
          reference_recovery_fraction = NA_real_,
          test_purity_fraction = NA_real_,
          eigengene_profile_cor = NA_real_,
          eigengene_profile_abs_cor = NA_real_,
          preservation_class = "not_matched"
        ))
      }

      test_module_label <- best_match$test_module_label[[1]]
      profile_cor <- NA_real_
      if (
        reference_module_label %in% names(reference_eigengenes) &&
          test_module_label %in% names(test_eigengenes)
      ) {
        profile_pair <- reference_eigengenes %>%
          dplyr::transmute(
            sample_code = .data$sample_code,
            reference_value = .data[[reference_module_label]]
          ) %>%
          dplyr::inner_join(
            test_eigengenes %>%
              dplyr::transmute(
                sample_code = .data$sample_code,
                test_value = .data[[test_module_label]]
              ),
            by = "sample_code"
          )

        profile_cor <- WGCNA::bicor(
          profile_pair$reference_value,
          profile_pair$test_value,
          use = "pairwise.complete.obs",
          maxPOutliers = max_p_outliers
        ) %>%
          as.numeric() %>%
          first()
      }

      preservation_class <- dplyr::case_when(
        is.finite(best_match$jaccard[[1]]) &&
          best_match$jaccard[[1]] >= 0.60 &&
          is.finite(profile_cor) &&
          abs(profile_cor) >= 0.75 ~ "high",
        is.finite(best_match$jaccard[[1]]) &&
          best_match$jaccard[[1]] >= 0.40 &&
          is.finite(profile_cor) &&
          abs(profile_cor) >= 0.60 ~ "moderate",
        TRUE ~ "low"
      )

      tibble::tibble(
        reference_network = reference_label,
        test_network = test_label,
        reference_size = reference_result$network_size,
        test_size = test_result$network_size,
        reference_module = reference_module,
        reference_module_label = reference_module_label,
        test_module = best_match$test_module[[1]],
        test_module_label = test_module_label,
        reference_module_shared_n = length(reference_gene_set),
        reference_to_test_grey_n = reference_to_grey_n,
        reference_to_test_grey_fraction = safe_divide(
          reference_to_grey_n,
          length(reference_gene_set)
        ),
        intersection_n = best_match$intersection_n[[1]],
        union_n = best_match$union_n[[1]],
        jaccard = best_match$jaccard[[1]],
        reference_recovery_fraction =
          best_match$reference_recovery_fraction[[1]],
        test_purity_fraction = best_match$test_purity_fraction[[1]],
        eigengene_profile_cor = profile_cor,
        eigengene_profile_abs_cor = abs(profile_cor),
        preservation_class = preservation_class
      )
    }
  )
}
```

# 1. Input manifest

```{r input-manifest}
vst_file <- file.path(
  project_dir,
  "Results_01_Global_Transcriptome_Structure",
  "matrices",
  "vst_matrix_germany.tsv"
)
metadata_file <- file.path(
  project_dir,
  "Results_01_Global_Transcriptome_Structure",
  "tables",
  "metadata_matched_germany.csv"
)
station_file <- file.path(
  project_dir,
  "01_metadata",
  "station_crosswalk.csv"
)
annotation_file <- file.path(
  project_dir,
  "Results_03_Differential_Expression",
  "tables",
  "gene_annotation_collapsed_from_trinotate.csv"
)
aggregate_deg_file <- file.path(
  project_dir,
  "Results_03_Differential_Expression",
  "tables",
  "reference_significant_deg_table.csv"
)
deg_source_tbl <- planned_contrasts %>%
  dplyr::transmute(
    contrast = .data$contrast,
    numerator_site = .data$numerator_site,
    denominator_site = .data$denominator_site,
    path = file.path(
      project_dir,
      "Results_03_Differential_Expression",
      "contrast_tables",
      .data$contrast,
      paste0(.data$contrast, "_significant_DEGs.csv")
    )
  )
gene_to_go_file <- file.path(
  project_dir,
  "Results_04_EchoGO_Seed_Preparation",
  "derived_inputs",
  "gene_to_go_primary_eggnog_long.csv"
)

input_manifest <- dplyr::bind_rows(tibble::tribble(
  ~input_id, ~path, ~required, ~role,
  "vst_expression", vst_file, TRUE,
  "gene-level VST expression matrix",
  "matched_metadata", metadata_file, TRUE,
  "sample and site metadata",
  "station_crosswalk", station_file, TRUE,
  "station order and site labels",
  "collapsed_annotation", annotation_file, FALSE,
  "gene annotation coverage",
  "aggregate_reference_degs", aggregate_deg_file, FALSE,
  "reference-only DEG aggregate retained for provenance; not used here",
  "primary_gene_to_go", gene_to_go_file, FALSE,
  "GO-mapping coverage"
), deg_source_tbl %>%
  dplyr::transmute(
    input_id = paste0("focal_degs_", .data$contrast),
    path = .data$path,
    required = TRUE,
    role = paste0(
      "required DEG source for ", .data$numerator_site,
      " versus ", .data$denominator_site
    )
  )) %>%
  dplyr::mutate(present = file.exists(.data$path))

write_csv_safe(
  input_manifest,
  file.path(tables_dir, "input_manifest.csv")
)

knitr::kable(
  input_manifest,
  caption = "Inputs for the WGCNA gene-set sensitivity analysis."
)

missing_required <- input_manifest %>%
  dplyr::filter(.data$required, !.data$present)

if (nrow(missing_required) > 0L) {
  stop(
    "Missing required inputs:\n",
    paste(missing_required$path, collapse = "\n")
  )
}
```

# 2. Load and rank the expression matrix

Genes are ranked once by variance across all 30 VST libraries. The 5,000-,
10,000-, and 15,000-gene networks are therefore nested and differ only by the
number of lower-ranked genes admitted to the network. Site labels are not used
for feature selection.

```{r load-and-rank-expression}
station_tbl <- readr::read_csv(
  station_file,
  show_col_types = FALSE
)
metadata_tbl <- readr::read_csv(
  metadata_file,
  show_col_types = FALSE
) %>%
  dplyr::mutate(
    sample_code = as.character(.data$sample_code),
    site_id = as.character(.data$site_id),
    station_tag = as.character(.data$station_tag)
  )

vst_tbl <- readr::read_tsv(
  vst_file,
  show_col_types = FALSE
)

sample_columns <- intersect(
  metadata_tbl$sample_code,
  setdiff(names(vst_tbl), "gene_id")
)

if (length(sample_columns) != nrow(metadata_tbl)) {
  stop(
    "Sample mismatch: ",
    length(sample_columns),
    " VST columns versus ",
    nrow(metadata_tbl),
    " metadata rows."
  )
}

metadata_aligned <- metadata_tbl %>%
  dplyr::slice(match(sample_columns, .data$sample_code)) %>%
  dplyr::left_join(
    station_tbl %>%
      dplyr::select(
        .data$legacy_site_id,
        dplyr::any_of(c(
          "station_order",
          "chemistry_series_order",
          "site_context",
          "chemistry_status"
        ))
      ),
    by = c("site_id" = "legacy_site_id")
  )

expression_tbl <- vst_tbl %>%
  dplyr::select(
    .data$gene_id,
    dplyr::all_of(sample_columns)
  ) %>%
  dplyr::distinct(.data$gene_id, .keep_all = TRUE)

expression_matrix <- as.matrix(
  expression_tbl[, sample_columns]
)
rownames(expression_matrix) <- expression_tbl$gene_id
storage.mode(expression_matrix) <- "double"

gene_variance <- matrixStats::rowVars(
  expression_matrix,
  na.rm = TRUE
)
names(gene_variance) <- rownames(expression_matrix)

valid_gene <- is.finite(gene_variance) & gene_variance > 0
expression_matrix <- expression_matrix[valid_gene, , drop = FALSE]
gene_variance <- gene_variance[valid_gene]
ranked_gene_ids <- names(
  sort(gene_variance, decreasing = TRUE)
)

if (length(ranked_gene_ids) < min(gene_set_sizes)) {
  stop(
    "Fewer than ",
    min(gene_set_sizes),
    " non-zero-variance genes are available."
  )
}

available_gene_set_sizes <- gene_set_sizes[
  gene_set_sizes <= length(ranked_gene_ids)
]
if (length(available_gene_set_sizes) == 0L) {
  stop("None of the requested gene-set sizes can be constructed.")
}
if (!all(gene_set_sizes %in% available_gene_set_sizes)) {
  warning(
    "Requested sizes exceeding the available non-zero-variance genes were skipped."
  )
}
gene_set_sizes <- available_gene_set_sizes

network_input_info <- file.info(c(vst_file, metadata_file, station_file))
network_input_signature <- paste(
  paste(
    basename(rownames(network_input_info)),
    network_input_info$size,
    as.character(unname(tools::md5sum(rownames(network_input_info)))),
    paste0("samples=", paste(sample_columns, collapse = ";"))
  ),
  collapse = "||"
)

expression_qc <- tibble::tibble(
  raw_vst_gene_n = nrow(vst_tbl),
  unique_gene_n = nrow(expression_tbl),
  nonzero_variance_gene_n = length(ranked_gene_ids),
  sample_n = length(sample_columns),
  tested_gene_sets = paste(gene_set_sizes, collapse = " | "),
  maximum_block_size = max_block_size
)

write_csv_safe(
  expression_qc,
  file.path(tables_dir, "expression_input_qc.csv")
)
write_csv_safe(
  tibble::tibble(
    gene_id = ranked_gene_ids,
    variance_rank = seq_along(ranked_gene_ids),
    vst_variance = unname(gene_variance[ranked_gene_ids])
  ),
  file.path(tables_dir, "vst_gene_variance_ranking.csv")
)

knitr::kable(
  expression_qc,
  caption = "Expression input and sensitivity-design summary."
)
```

# 3. Reconstruct the networks

Each network uses the same signed biweight-midcorrelation design. The
soft-thresholding power is selected independently using the current Step 02 rule:
the lowest tested power reaching the target is retained; when the target is not
reached, the power with the highest available fit is used. Diagnostics at power
12 are also retained so that the effect of adding genes can be compared at a
common power.

```{r reconstruct-networks, message=TRUE}
network_results <- purrr::map(
  gene_set_sizes,
  load_or_run_network_for_gene_set,
  ranked_gene_ids = ranked_gene_ids,
  expression_matrix = expression_matrix,
  metadata_aligned = metadata_aligned,
  input_signature = network_input_signature
)
names(network_results) <- vapply(
  network_results,
  function(current_result) current_result$network_label,
  character(1)
)

network_summary <- purrr::map_dfr(
  network_results,
  "summary"
) %>%
  dplyr::arrange(.data$requested_gene_n)

soft_threshold_all <- purrr::map_dfr(
  network_results,
  "soft"
)
module_summary_all <- purrr::map_dfr(
  network_results,
  "modules"
)
module_assignment_all <- purrr::map_dfr(
  network_results,
  "assignment"
)
planned_contrasts_all <- purrr::map_dfr(
  network_results,
  "contrasts"
)

write_csv_safe(
  network_summary,
  file.path(tables_dir, "network_size_diagnostic_summary.csv")
)
write_csv_safe(
  soft_threshold_all,
  file.path(tables_dir, "soft_threshold_diagnostics_all_gene_sets.csv")
)
write_csv_safe(
  module_summary_all,
  file.path(tables_dir, "module_summary_all_gene_sets.csv")
)
write_csv_safe(
  planned_contrasts_all,
  file.path(tables_dir, "planned_module_contrasts_all_gene_sets.csv")
)

knitr::kable(
  network_summary,
  digits = 4,
  caption = "WGCNA diagnostics across the tested gene-set sizes."
)
```

# 4. Soft-threshold and connectivity diagnostics

```{r soft-threshold-comparison, fig.width=11, fig.height=7.5}
soft_threshold_plot_tbl <- soft_threshold_all %>%
  dplyr::select(
    .data$network_label,
    .data$requested_gene_n,
    .data$power,
    .data$scale_free_fit,
    .data$mean_connectivity
  ) %>%
  tidyr::pivot_longer(
    cols = c(
      .data$scale_free_fit,
      .data$mean_connectivity
    ),
    names_to = "metric",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    metric = dplyr::recode(
      .data$metric,
      scale_free_fit = "Scale-free topology fit",
      mean_connectivity = "Mean connectivity"
    ),
    gene_set = paste0(
      format(.data$requested_gene_n, big.mark = ","),
      " genes"
    )
  )

soft_threshold_plot <- ggplot2::ggplot(
  soft_threshold_plot_tbl,
  ggplot2::aes(
    x = .data$power,
    y = .data$value,
    group = .data$gene_set,
    linetype = .data$gene_set,
    shape = .data$gene_set
  )
) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::geom_point(size = 2.1) +
  ggplot2::facet_wrap(
    ~metric,
    scales = "free_y",
    ncol = 1
  ) +
  ggplot2::labs(
    title = "Soft-threshold diagnostics across WGCNA gene-set sizes",
    subtitle = paste(
      "Power is re-evaluated for each network;",
      "the target fit is 0.80"
    ),
    x = "Soft-thresholding power",
    y = "Diagnostic value",
    linetype = "Gene set",
    shape = "Gene set"
  ) +
  theme_publication()

save_plot_bundle(
  soft_threshold_plot,
  file.path(plots_dir, "soft_threshold_gene_set_comparison"),
  width = 11,
  height = 7.5
)

soft_threshold_plot
```

```{r common-power-comparison}
common_power_summary <- soft_threshold_all %>%
  dplyr::filter(
    .data$power == reference_power_for_comparison
  ) %>%
  dplyr::select(
    .data$network_label,
    .data$requested_gene_n,
    .data$power,
    .data$scale_free_fit,
    .data$slope,
    .data$mean_connectivity,
    .data$median_connectivity,
    .data$max_connectivity
  ) %>%
  dplyr::arrange(.data$requested_gene_n)

write_csv_safe(
  common_power_summary,
  file.path(
    tables_dir,
    paste0(
      "common_power_",
      reference_power_for_comparison,
      "_diagnostics.csv"
    )
  )
)

knitr::kable(
  common_power_summary,
  digits = 4,
  caption = paste(
    "Network diagnostics at the common soft-thresholding power",
    reference_power_for_comparison,
    "."
  )
)
```

```{r network-summary-plot, fig.width=12, fig.height=6.5}
network_metric_plot_tbl <- network_summary %>%
  dplyr::select(
    .data$network_label,
    .data$requested_gene_n,
    .data$selected_scale_free_fit,
    .data$selected_mean_connectivity,
    .data$assigned_gene_fraction,
    .data$n_assigned_modules,
    .data$grey_gene_fraction
  ) %>%
  tidyr::pivot_longer(
    cols = -c(
      .data$network_label,
      .data$requested_gene_n
    ),
    names_to = "metric",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    metric = dplyr::recode(
      .data$metric,
      selected_scale_free_fit = "Selected-power scale-free fit",
      selected_mean_connectivity = "Selected-power mean connectivity",
      assigned_gene_fraction = "Assigned gene fraction",
      n_assigned_modules = "Assigned modules",
      grey_gene_fraction = "Grey gene fraction"
    ),
    gene_set = factor(
      paste0(
        format(.data$requested_gene_n, big.mark = ","),
        " genes"
      ),
      levels = paste0(
        format(sort(unique(.data$requested_gene_n)), big.mark = ","),
        " genes"
      )
    )
  )

network_metric_plot <- ggplot2::ggplot(
  network_metric_plot_tbl,
  ggplot2::aes(
    x = .data$gene_set,
    y = .data$value,
    group = 1
  )
) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::geom_point(size = 2.8) +
  ggplot2::facet_wrap(
    ~metric,
    scales = "free_y",
    ncol = 3
  ) +
  ggplot2::labs(
    title = "Network structure across WGCNA gene-set sizes",
    x = NULL,
    y = "Value"
  ) +
  theme_publication(base_size = 10) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
  )

save_plot_bundle(
  network_metric_plot,
  file.path(plots_dir, "network_metric_gene_set_comparison"),
  width = 12,
  height = 6.5
)

network_metric_plot
```

# 5. Preservation of modules across nested gene sets

Reference modules are matched to modules in the larger network after restricting
both assignments to their shared genes. Jaccard similarity measures overlap in
gene membership. Eigengene-profile correlation measures whether the matched
modules retain the same dominant expression pattern across the 30 libraries;
its absolute value is reported because eigengene signs are arbitrary. The
high/moderate/low classes are transparent heuristics for screening and are not
formal WGCNA preservation statistics.

```{r module-preservation-across-sizes}
network_pairs <- tibble::tribble(
  ~reference_size, ~test_size,
  5000L, 10000L,
  5000L, 15000L,
  10000L, 15000L
) %>%
  dplyr::filter(
    .data$reference_size %in% gene_set_sizes,
    .data$test_size %in% gene_set_sizes
  )

module_preservation <- purrr::pmap_dfr(
  network_pairs,
  function(reference_size, test_size) {
    reference_result <- network_results[[
      paste0("G", reference_size)
    ]]
    test_result <- network_results[[
      paste0("G", test_size)
    ]]

    compare_network_pair(
      reference_result,
      test_result
    )
  }
)

preservation_summary <- module_preservation %>%
  dplyr::group_by(
    .data$reference_network,
    .data$test_network,
    .data$reference_size,
    .data$test_size
  ) %>%
  dplyr::summarise(
    n_reference_modules = dplyr::n(),
    median_jaccard = stats::median(
      .data$jaccard,
      na.rm = TRUE
    ),
    minimum_jaccard = min(
      .data$jaccard,
      na.rm = TRUE
    ),
    median_reference_recovery = stats::median(
      .data$reference_recovery_fraction,
      na.rm = TRUE
    ),
    median_abs_eigengene_cor = stats::median(
      .data$eigengene_profile_abs_cor,
      na.rm = TRUE
    ),
    n_high_preservation = sum(
      .data$preservation_class == "high",
      na.rm = TRUE
    ),
    n_moderate_preservation = sum(
      .data$preservation_class == "moderate",
      na.rm = TRUE
    ),
    n_low_preservation = sum(
      .data$preservation_class == "low",
      na.rm = TRUE
    ),
    .groups = "drop"
  )

write_csv_safe(
  module_preservation,
  file.path(tables_dir, "module_preservation_across_gene_sets.csv")
)
write_csv_safe(
  preservation_summary,
  file.path(tables_dir, "module_preservation_summary.csv")
)

knitr::kable(
  preservation_summary,
  digits = 4,
  caption = "Module preservation across nested WGCNA gene sets."
)

knitr::kable(
  module_preservation %>%
    dplyr::arrange(
      .data$reference_size,
      .data$test_size,
      dplyr::desc(.data$jaccard)
    ),
  digits = 4,
  caption = "Best module matches based on shared-gene Jaccard similarity."
)
```

```{r module-preservation-plot, fig.width=11, fig.height=7}
preservation_plot_tbl <- module_preservation %>%
  dplyr::mutate(
    comparison = paste0(
      format(.data$reference_size, big.mark = ","),
      " -> ",
      format(.data$test_size, big.mark = ",")
    )
  )

module_preservation_plot <- ggplot2::ggplot(
  preservation_plot_tbl,
  ggplot2::aes(
    x = .data$jaccard,
    y = .data$eigengene_profile_abs_cor,
    shape = .data$preservation_class
  )
) +
  ggplot2::geom_vline(
    xintercept = c(0.40, 0.60),
    linetype = "dashed",
    linewidth = 0.4
  ) +
  ggplot2::geom_hline(
    yintercept = c(0.60, 0.75),
    linetype = "dashed",
    linewidth = 0.4
  ) +
  ggplot2::geom_point(size = 2.8, alpha = 0.85) +
  ggplot2::facet_wrap(~comparison) +
  ggplot2::labs(
    title = "Preservation of module membership and eigengene profiles",
    subtitle = paste(
      "Each point is one reference module matched to its best module",
      "in the larger gene set"
    ),
    x = "Shared-gene Jaccard similarity",
    y = "Absolute eigengene-profile correlation",
    shape = "Heuristic class"
  ) +
  theme_publication()

save_plot_bundle(
  module_preservation_plot,
  file.path(plots_dir, "module_preservation_gene_set_comparison"),
  width = 11,
  height = 7
)

module_preservation_plot
```

# 6. Preservation of planned site-contrast signals

The three planned contrasts are fitted separately within each reconstructed
network using the individual-library module eigengenes. Benjamini-Hochberg
correction is applied across all module-by-contrast tests within each network.
Because module labels and eigengene signs are arbitrary across networks, direct
cross-network effect comparisons are made only for the best-matched modules and
the test-network effect is aligned to the sign of the eigengene-profile
correlation.

```{r contrast-preservation}
contrast_preservation <- module_preservation %>%
  dplyr::filter(!is.na(.data$test_module_label)) %>%
  dplyr::select(
    .data$reference_network,
    .data$test_network,
    .data$reference_size,
    .data$test_size,
    .data$reference_module_label,
    .data$test_module_label,
    .data$jaccard,
    .data$eigengene_profile_cor,
    .data$eigengene_profile_abs_cor,
    .data$preservation_class
  ) %>%
  dplyr::left_join(
    planned_contrasts_all %>%
      dplyr::select(
        reference_network = .data$network_label,
        reference_module_label = .data$module_label,
        .data$contrast,
        .data$contrast_panel,
        reference_hedges_g = .data$hedges_g,
        reference_p = .data$p_value,
        reference_q = .data$q_global_network,
        reference_fdr_supported = .data$fdr_supported,
        reference_large_nominal = .data$large_nominal_effect
      ),
    by = c(
      "reference_network",
      "reference_module_label"
    )
  ) %>%
  dplyr::left_join(
    planned_contrasts_all %>%
      dplyr::select(
        test_network = .data$network_label,
        test_module_label = .data$module_label,
        .data$contrast,
        test_hedges_g = .data$hedges_g,
        test_p = .data$p_value,
        test_q = .data$q_global_network,
        test_fdr_supported = .data$fdr_supported,
        test_large_nominal = .data$large_nominal_effect
      ),
    by = c(
      "test_network",
      "test_module_label",
      "contrast"
    )
  ) %>%
  dplyr::mutate(
    alignment_sign = dplyr::if_else(
      is.finite(.data$eigengene_profile_cor) &
        .data$eigengene_profile_cor < 0,
      -1,
      1
    ),
    test_hedges_g_aligned =
      .data$test_hedges_g * .data$alignment_sign,
    aligned_effect_difference =
      .data$test_hedges_g_aligned - .data$reference_hedges_g,
    same_aligned_direction = dplyr::case_when(
      !is.finite(.data$reference_hedges_g) |
        !is.finite(.data$test_hedges_g_aligned) ~ NA,
      .data$reference_hedges_g == 0 |
        .data$test_hedges_g_aligned == 0 ~ NA,
      TRUE ~ sign(.data$reference_hedges_g) ==
        sign(.data$test_hedges_g_aligned)
    )
  )

contrast_network_summary <- planned_contrasts_all %>%
  dplyr::group_by(
    .data$network_label,
    .data$network_size,
    .data$contrast,
    .data$contrast_panel
  ) %>%
  dplyr::summarise(
    n_tested_modules = dplyr::n(),
    n_fdr_supported = sum(
      .data$fdr_supported,
      na.rm = TRUE
    ),
    n_large_nominal = sum(
      .data$large_nominal_effect,
      na.rm = TRUE
    ),
    maximum_abs_hedges_g = max(
      abs(.data$hedges_g),
      na.rm = TRUE
    ),
    minimum_global_q = min(
      .data$q_global_network,
      na.rm = TRUE
    ),
    leading_module = .data$module_label[
      which.max(abs(.data$hedges_g))
    ][[1]],
    .groups = "drop"
  )

write_csv_safe(
  contrast_preservation,
  file.path(tables_dir, "planned_contrast_preservation_across_gene_sets.csv")
)
write_csv_safe(
  contrast_network_summary,
  file.path(tables_dir, "planned_contrast_summary_by_gene_set.csv")
)

knitr::kable(
  contrast_network_summary,
  digits = 4,
  caption = "Planned module-contrast evidence across gene-set sizes."
)
```

```{r planned-contrast-heatmap, fig.width=13, fig.height=10}
contrast_heatmap_tbl <- planned_contrasts_all %>%
  dplyr::mutate(
    gene_set = factor(
      paste0(
        format(.data$network_size, big.mark = ","),
        " genes"
      ),
      levels = paste0(
        format(sort(unique(.data$network_size)), big.mark = ","),
        " genes"
      )
    ),
    module_label = factor(
      .data$module_label,
      levels = rev(unique(.data$module_label))
    ),
    cell_label = dplyr::case_when(
      .data$q_global_network < 0.05 ~
        paste0("g=", sprintf("%.2f", .data$hedges_g), "\nq<0.05"),
      .data$p_value < 0.05 ~
        paste0("g=", sprintf("%.2f", .data$hedges_g), "\np<0.05"),
      TRUE ~ paste0("g=", sprintf("%.2f", .data$hedges_g))
    )
  )

planned_contrast_heatmap <- ggplot2::ggplot(
  contrast_heatmap_tbl,
  ggplot2::aes(
    x = .data$contrast_panel,
    y = .data$module_label,
    fill = .data$hedges_g
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.35) +
  ggplot2::geom_text(
    ggplot2::aes(label = .data$cell_label),
    size = 2.5
  ) +
  ggplot2::scale_fill_gradient2(
    low = "#2c7bb6",
    mid = "white",
    high = "#d7191c",
    midpoint = 0,
    name = "Hedges g"
  ) +
  ggplot2::facet_wrap(
    ~gene_set,
    scales = "free_y",
    ncol = 1
  ) +
  ggplot2::labs(
    title = "Planned module-eigengene contrasts across gene-set sizes",
    subtitle = paste(
      "Module labels are network-specific; compare matched modules using",
      "the preservation tables"
    ),
    x = NULL,
    y = "Network-specific module"
  ) +
  theme_publication(base_size = 9) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 20, hjust = 1)
  )

save_plot_bundle(
  planned_contrast_heatmap,
  file.path(plots_dir, "planned_contrast_heatmap_all_gene_sets"),
  width = 13,
  height = 10
)

planned_contrast_heatmap
```

# 7. DEG and annotation coverage

Increasing the gene set can improve biological coverage even when the
scale-free-topology fit does not improve. This section records how many focal
DEGs, annotated genes, and GO-mapped genes are eligible for the network and how
many are retained in assigned modules. These diagnostics do not use DEG status
or annotation to construct the network.

```{r load-deg-and-annotation-coverage}
focal_deg_objects <- purrr::pmap(
  deg_source_tbl,
  function(contrast, numerator_site, denominator_site, path) {
    if (!file.exists(path)) {
      stop("Required DEG file is missing for ", contrast, ": ", path)
    }
    raw_tbl <- readr::read_csv(path, show_col_types = FALSE)
    required_columns <- c(
      "gene_id", "contrast", "numerator", "denominator", "padj",
      "log2FoldChange_shrunk"
    )
    missing_columns <- setdiff(required_columns, names(raw_tbl))
    if (length(missing_columns) > 0L) {
      stop(
        "DEG file for ", contrast, " is missing: ",
        paste(missing_columns, collapse = ", ")
      )
    }
    observed_contrasts <- unique(as.character(raw_tbl$contrast))
    observed_numerators <- unique(as.character(raw_tbl$numerator))
    observed_denominators <- unique(as.character(raw_tbl$denominator))
    if (!identical(observed_contrasts, contrast) ||
        !identical(observed_numerators, numerator_site) ||
        !identical(observed_denominators, denominator_site)) {
      stop(
        "DEG direction audit failed for ", contrast,
        ". Expected ", numerator_site, " versus ", denominator_site, "."
      )
    }

    filtered_tbl <- raw_tbl %>%
      dplyr::filter(
        is.finite(.data$padj),
        .data$padj < 0.05,
        is.finite(.data$log2FoldChange_shrunk),
        abs(.data$log2FoldChange_shrunk) >= 1
      ) %>%
      dplyr::transmute(
        gene_id = as.character(.data$gene_id),
        contrast = as.character(.data$contrast),
        numerator = as.character(.data$numerator),
        denominator = as.character(.data$denominator),
        padj = as.numeric(.data$padj),
        log2FoldChange_shrunk = as.numeric(.data$log2FoldChange_shrunk),
        direction = dplyr::if_else(
          .data$log2FoldChange_shrunk > 0,
          "higher_in_numerator",
          "lower_in_numerator"
        ),
        source_file = normalizePath(path, winslash = "/", mustWork = TRUE)
      ) %>%
      dplyr::distinct(.data$gene_id, .keep_all = TRUE)

    list(
      genes = filtered_tbl,
      audit = tibble::tibble(
        contrast = contrast,
        numerator = numerator_site,
        denominator = denominator_site,
        source_file = normalizePath(path, winslash = "/", mustWork = TRUE),
        input_row_n = nrow(raw_tbl),
        padj_lt_0_05_n = sum(
          is.finite(raw_tbl$padj) & raw_tbl$padj < 0.05,
          na.rm = TRUE
        ),
        abs_shrunken_log2fc_ge_1_n = sum(
          is.finite(raw_tbl$log2FoldChange_shrunk) &
            abs(raw_tbl$log2FoldChange_shrunk) >= 1,
          na.rm = TRUE
        ),
        final_deg_n = nrow(filtered_tbl),
        criteria = "padj < 0.05 and abs(log2FoldChange_shrunk) >= 1"
      )
    )
  }
)

focal_deg_tbl <- purrr::map_dfr(focal_deg_objects, "genes")
deg_input_audit <- purrr::map_dfr(focal_deg_objects, "audit")

missing_contrasts <- setdiff(
  planned_contrasts$contrast,
  unique(focal_deg_tbl$contrast)
)
if (length(missing_contrasts) > 0L) {
  stop("No qualifying DEGs were loaded for: ", paste(missing_contrasts, collapse = ", "))
}
if (deg_input_audit$final_deg_n[
      deg_input_audit$contrast == "HOL05_vs_HOL04"
    ] == 0L) {
  stop("H6-H5 DEG loading returned zero genes after the declared filters.")
}

write_csv_safe(
  deg_input_audit,
  file.path(tables_dir, "focal_deg_input_audit.csv")
)

annotation_gene_tbl <- tibble::tibble(
  gene_id = character(),
  has_readable_annotation = logical()
)

if (file.exists(annotation_file)) {
  raw_annotation_tbl <- readr::read_csv(
    annotation_file,
    show_col_types = FALSE
  )

  if ("gene_id" %in% names(raw_annotation_tbl)) {
    candidate_annotation_columns <- intersect(
      c(
        "best_gene_label",
        "gene_label",
        "EggNM.Preferred_name",
        "EggNM.Description",
        "consensus_gene_symbol",
        "annotation"
      ),
      names(raw_annotation_tbl)
    )

    if (length(candidate_annotation_columns) > 0L) {
      annotation_values <- dplyr::select(
        raw_annotation_tbl,
        dplyr::all_of(candidate_annotation_columns)
      )
      annotation_gene_ids <- as.character(raw_annotation_tbl$gene_id)
      has_readable_annotation <- vapply(
        seq_len(nrow(raw_annotation_tbl)),
        function(row_index) {
          current_values <- safe_chr(
            unlist(annotation_values[row_index, ], use.names = FALSE)
          )
          current_values <- current_values[
            !is.na(current_values) &
              !current_values %in% c("NA", ".", "-") &
              current_values != annotation_gene_ids[[row_index]] &
              !stringr::str_detect(current_values, "^TRINITY_")
          ]
          length(current_values) > 0L
        },
        logical(1)
      )

      annotation_gene_tbl <- tibble::tibble(
        gene_id = annotation_gene_ids,
        has_readable_annotation = has_readable_annotation
      ) %>%
        dplyr::group_by(.data$gene_id) %>%
        dplyr::summarise(
          has_readable_annotation = any(
            .data$has_readable_annotation,
            na.rm = TRUE
          ),
          .groups = "drop"
        )
    }
  }
}

go_gene_ids <- character()
if (file.exists(gene_to_go_file)) {
  gene_to_go_tbl <- readr::read_csv(
    gene_to_go_file,
    show_col_types = FALSE
  )
  if ("gene_id" %in% names(gene_to_go_tbl)) {
    go_gene_ids <- unique(as.character(gene_to_go_tbl$gene_id))
  }
}

network_gene_coverage <- purrr::map_dfr(
  network_results,
  function(current_result) {
    current_assignment <- current_result$assignment
    eligible_gene_ids <- unique(current_assignment$gene_id)
    assigned_gene_ids <- unique(
      current_assignment$gene_id[
        current_assignment$module != "grey"
      ]
    )

    annotated_gene_ids <- annotation_gene_tbl$gene_id[
      annotation_gene_tbl$has_readable_annotation
    ]

    tibble::tibble(
      network_label = current_result$network_label,
      network_size = current_result$network_size,
      eligible_gene_n = length(eligible_gene_ids),
      assigned_gene_n = length(assigned_gene_ids),
      readable_annotation_eligible_n = length(intersect(
        eligible_gene_ids,
        annotated_gene_ids
      )),
      readable_annotation_assigned_n = length(intersect(
        assigned_gene_ids,
        annotated_gene_ids
      )),
      go_mapped_eligible_n = length(intersect(
        eligible_gene_ids,
        go_gene_ids
      )),
      go_mapped_assigned_n = length(intersect(
        assigned_gene_ids,
        go_gene_ids
      ))
    ) %>%
      dplyr::mutate(
        readable_annotation_eligible_fraction = safe_divide(
          .data$readable_annotation_eligible_n,
          .data$eligible_gene_n
        ),
        readable_annotation_assigned_fraction = safe_divide(
          .data$readable_annotation_assigned_n,
          .data$assigned_gene_n
        ),
        go_mapped_eligible_fraction = safe_divide(
          .data$go_mapped_eligible_n,
          .data$eligible_gene_n
        ),
        go_mapped_assigned_fraction = safe_divide(
          .data$go_mapped_assigned_n,
          .data$assigned_gene_n
        )
      )
  }
)

deg_coverage <- purrr::map_dfr(
  network_results,
  function(current_result) {
    current_assignment <- current_result$assignment
    eligible_gene_ids <- unique(current_assignment$gene_id)
    assigned_gene_ids <- unique(
      current_assignment$gene_id[
        current_assignment$module != "grey"
      ]
    )

    planned_contrasts %>%
      dplyr::select(
        .data$contrast,
        .data$contrast_panel
      ) %>%
      dplyr::mutate(
        network_label = current_result$network_label,
        network_size = current_result$network_size,
        total_significant_deg_n = purrr::map_int(
          .data$contrast,
          function(current_contrast) {
            dplyr::n_distinct(
              focal_deg_tbl$gene_id[
                focal_deg_tbl$contrast == current_contrast
              ]
            )
          }
        ),
        eligible_significant_deg_n = purrr::map_int(
          .data$contrast,
          function(current_contrast) {
            current_degs <- focal_deg_tbl$gene_id[
              focal_deg_tbl$contrast == current_contrast
            ]
            length(intersect(current_degs, eligible_gene_ids))
          }
        ),
        assigned_significant_deg_n = purrr::map_int(
          .data$contrast,
          function(current_contrast) {
            current_degs <- focal_deg_tbl$gene_id[
              focal_deg_tbl$contrast == current_contrast
            ]
            length(intersect(current_degs, assigned_gene_ids))
          }
        )
      ) %>%
      dplyr::mutate(
        eligible_deg_fraction = safe_divide(
          .data$eligible_significant_deg_n,
          .data$total_significant_deg_n
        ),
        assigned_deg_fraction_of_total = safe_divide(
          .data$assigned_significant_deg_n,
          .data$total_significant_deg_n
        ),
        assigned_deg_fraction_of_eligible = safe_divide(
          .data$assigned_significant_deg_n,
          .data$eligible_significant_deg_n
        )
      )
  }
)

write_csv_safe(
  network_gene_coverage,
  file.path(tables_dir, "annotation_and_go_coverage_by_gene_set.csv")
)
write_csv_safe(
  deg_coverage,
  file.path(tables_dir, "focal_deg_coverage_by_gene_set.csv")
)

knitr::kable(
  network_gene_coverage,
  digits = 4,
  caption = "Readable annotation and GO-mapping coverage across gene sets."
)

if (nrow(focal_deg_tbl) > 0L) {
  knitr::kable(
    deg_coverage,
    digits = 4,
    caption = "Coverage of significant genes from the three focal contrasts."
  )
} else {
  cat(
    "The aggregate focal DEG table was not available; DEG coverage was not evaluated."
  )
}
```

```{r deg-coverage-plot, fig.width=11, fig.height=6}
if (nrow(focal_deg_tbl) > 0L) {
  deg_coverage_plot_tbl <- deg_coverage %>%
    dplyr::select(
      .data$network_size,
      .data$contrast_panel,
      .data$eligible_deg_fraction,
      .data$assigned_deg_fraction_of_total
    ) %>%
    tidyr::pivot_longer(
      cols = c(
        .data$eligible_deg_fraction,
        .data$assigned_deg_fraction_of_total
      ),
      names_to = "coverage_type",
      values_to = "fraction"
    ) %>%
    dplyr::mutate(
      coverage_type = dplyr::recode(
        .data$coverage_type,
        eligible_deg_fraction = "Eligible for network",
        assigned_deg_fraction_of_total = "Assigned to a module"
      ),
      gene_set = factor(
        paste0(
          format(.data$network_size, big.mark = ","),
          " genes"
        ),
        levels = paste0(
          format(sort(unique(.data$network_size)), big.mark = ","),
          " genes"
        )
      )
    )

  deg_coverage_plot <- ggplot2::ggplot(
    deg_coverage_plot_tbl,
    ggplot2::aes(
      x = .data$gene_set,
      y = .data$fraction,
      group = .data$coverage_type,
      linetype = .data$coverage_type,
      shape = .data$coverage_type
    )
  ) +
    ggplot2::geom_line(linewidth = 0.7) +
    ggplot2::geom_point(size = 2.6) +
    ggplot2::facet_wrap(~contrast_panel) +
    ggplot2::scale_y_continuous(
      labels = scales::percent_format(accuracy = 1)
    ) +
    ggplot2::labs(
      title = "Recovery of focal DEGs across WGCNA gene-set sizes",
      x = NULL,
      y = "Fraction of significant DEGs",
      linetype = "Coverage",
      shape = "Coverage"
    ) +
    theme_publication(base_size = 10) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
    )

  save_plot_bundle(
    deg_coverage_plot,
    file.path(plots_dir, "focal_deg_coverage_gene_set_comparison"),
    width = 11,
    height = 6
  )

  deg_coverage_plot
}
```

# 8. Structural power sensitivity for the 10,000-gene candidate

The 10,000-gene candidate is reconstructed at powers 12, 14, 16, 18, and 20
with every other network setting fixed. Selection uses connectivity, module
assignment, adjacent-power preservation, and site-stratified resampling only.
Site contrasts, DEG overlap, enrichment yield, and chemistry are not available
to the power-selection rule.

```{r ten-thousand-power-sensitivity, message=TRUE}
g10000_gene_ids <- utils::head(ranked_gene_ids, 10000L)
dat_expr_g10000 <- t(
  expression_matrix[g10000_gene_ids, , drop = FALSE]
)
rownames(dat_expr_g10000) <- metadata_aligned$sample_code

g10000_soft_tbl <- network_results$G10000$soft %>%
  dplyr::filter(.data$power %in% power_sensitivity_candidates)

connectivity_cache_signature <- paste(
  network_input_signature,
  paste0("gene_hash=", digest::digest(g10000_gene_ids, algo = "xxhash64")),
  paste0("powers=", paste(power_sensitivity_candidates, collapse = ";")),
  paste0("network_type=", network_type),
  paste0("correlation_type=", correlation_type),
  paste0("max_p_outliers=", max_p_outliers),
  paste0("very_low_threshold=", very_low_connectivity_threshold),
  sep = "||"
)
connectivity_cache_file <- file.path(
  objects_dir,
  paste0(
    "power_connectivity_distribution_",
    digest::digest(connectivity_cache_signature, algo = "xxhash64"),
    ".rds"
  )
)
if (reuse_power_sensitivity_cache && file.exists(connectivity_cache_file)) {
  cached_connectivity <- readRDS(connectivity_cache_file)
  if (is.list(cached_connectivity) &&
      identical(cached_connectivity$signature, connectivity_cache_signature)) {
    power_connectivity_distribution <- cached_connectivity$result
  }
}
if (!exists("power_connectivity_distribution", inherits = FALSE)) {
  power_connectivity_distribution <- purrr::map_dfr(
    power_sensitivity_candidates,
    function(current_power) {
      message("Calculating connectivity distribution at power ", current_power)
      connectivity <- WGCNA::softConnectivity(
        dat_expr_g10000,
        corFnc = correlation_type,
        corOptions = paste0(
          "use = 'p', maxPOutliers = ", max_p_outliers
        ),
        type = network_type,
        power = current_power,
        blockSize = 1000,
        verbose = 0
      )
      tibble::tibble(
        power = current_power,
        connectivity_p10 = unname(stats::quantile(
          connectivity, 0.10, na.rm = TRUE
        )),
        connectivity_p90 = unname(stats::quantile(
          connectivity, 0.90, na.rm = TRUE
        )),
        proportion_very_low_connectivity = mean(
          connectivity < very_low_connectivity_threshold,
          na.rm = TRUE
        ),
        very_low_connectivity_threshold = very_low_connectivity_threshold
      )
    }
  )
  saveRDS(
    list(
      signature = connectivity_cache_signature,
      result = power_connectivity_distribution
    ),
    connectivity_cache_file,
    compress = FALSE
  )
}

power_cache_signature_base <- paste(
  network_input_signature,
  paste0(
    "gene_hash=",
    digest::digest(g10000_gene_ids, algo = "xxhash64")
  ),
  paste0("WGCNA=", as.character(utils::packageVersion("WGCNA"))),
  paste0("seed=", power_sensitivity_seed),
  paste0("network_type=", network_type),
  paste0("correlation_type=", correlation_type),
  paste0("max_p_outliers=", max_p_outliers),
  paste0("minimum_module_size=", minimum_module_size),
  paste0("merge_cut_height=", merge_cut_height),
  paste0("deep_split=", deep_split),
  paste0("pam_respects_dendro=", pam_respects_dendro),
  paste0("max_block_size=", max_block_size),
  "undersized_module_policy=reclassify_to_exact_grey_v1",
  "eigengene_orientation=positive_average_standardized_expression_v1",
  sep = "||"
)

power_network_cache_dir <- file.path(
  objects_dir, "power_sensitivity_network_cache"
)
power_resample_cache_dir <- file.path(
  objects_dir, "power_sensitivity_resample_cache"
)
dir.create(power_network_cache_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(power_resample_cache_dir, recursive = TRUE, showWarnings = FALSE)

load_or_run_power_network <- function(current_power) {
  current_signature <- paste(
    power_cache_signature_base,
    paste0("power=", current_power),
    sep = "||"
  )
  cache_file <- file.path(
    power_network_cache_dir,
    paste0(
      "G10000_power_", current_power, "_",
      digest::digest(current_signature, algo = "xxhash64"),
      ".rds"
    )
  )
  if (reuse_power_sensitivity_cache && file.exists(cache_file)) {
    cached <- readRDS(cache_file)
    if (is.list(cached) &&
        identical(cached$signature, current_signature)) {
      message("Reusing 10,000-gene power cache: ", current_power)
      return(cached$result)
    }
  }
  set.seed(power_sensitivity_seed + current_power)
  result <- wgcna_run_validated_network(
    dat_expr = dat_expr_g10000,
    power = current_power,
    network_type = network_type,
    correlation_type = correlation_type,
    max_p_outliers = max_p_outliers,
    minimum_module_size = minimum_module_size,
    merge_cut_height = merge_cut_height,
    deep_split = deep_split,
    pam_respects_dendro = pam_respects_dendro,
    max_block_size = max_block_size,
    verbose = 0
  )
  saveRDS(
    list(signature = current_signature, result = result),
    cache_file,
    compress = FALSE
  )
  result
}

power_network_results <- purrr::map(
  power_sensitivity_candidates,
  load_or_run_power_network
)
names(power_network_results) <- paste0("P", power_sensitivity_candidates)

power_module_audit <- purrr::imap_dfr(
  power_network_results,
  function(current_result, current_label) {
    current_result$module_audit %>%
      dplyr::mutate(
        power = as.integer(stringr::str_remove(current_label, "^P")),
        .before = 1
      )
  }
)
power_anomalous_gene_audit <- purrr::imap_dfr(
  power_network_results,
  function(current_result, current_label) {
    current_result$anomalous_gene_audit %>%
      dplyr::mutate(
        power = as.integer(stringr::str_remove(current_label, "^P")),
        .before = 1
      )
  }
)
power_eigengene_orientation <- purrr::imap_dfr(
  power_network_results,
  function(current_result, current_label) {
    current_result$orientation %>%
      dplyr::mutate(
        power = as.integer(stringr::str_remove(current_label, "^P")),
        .before = 1
      )
  }
)

power_network_structure <- purrr::imap_dfr(
  power_network_results,
  function(current_result, current_label) {
    current_power <- as.integer(stringr::str_remove(current_label, "^P"))
    assigned_sizes <- current_result$module_summary$n_genes[
      current_result$module_summary$module != "grey"
    ]
    grey_n <- sum(current_result$module_summary$n_genes[
      current_result$module_summary$module == "grey"
    ])
    tibble::tibble(
      power = current_power,
      assigned_gene_fraction = (10000L - grey_n) / 10000L,
      grey_fraction = grey_n / 10000L,
      module_n = length(assigned_sizes),
      minimum_module_size_observed = min(assigned_sizes),
      median_module_size = stats::median(assigned_sizes),
      maximum_module_size = max(assigned_sizes),
      n_blocks = current_result$n_blocks,
      raw_undersized_module_n = sum(
        current_result$module_audit$below_declared_minimum
      )
    )
  }
)

adjacent_power_pairs <- tibble::tibble(
  reference_power = power_sensitivity_candidates[-length(power_sensitivity_candidates)],
  alternative_power = power_sensitivity_candidates[-1]
)
power_adjacent_preservation <- purrr::pmap_dfr(
  adjacent_power_pairs,
  function(reference_power, alternative_power) {
    reference_result <- power_network_results[[paste0("P", reference_power)]]
    alternative_result <- power_network_results[[paste0("P", alternative_power)]]
    wgcna_match_modules(
      reference_colors = reference_result$colors,
      alternative_colors = alternative_result$colors,
      reference_eigengenes = reference_result$eigengenes,
      projected_alternative_eigengenes = alternative_result$eigengenes,
      max_p_outliers = max_p_outliers
    ) %>%
      dplyr::mutate(
        reference_power = reference_power,
        alternative_power = alternative_power,
        .before = 1
      )
  }
)

set.seed(power_sensitivity_seed)
power_resample_design <- purrr::map_dfr(
  seq_len(power_resample_replicates),
  function(resample_id) {
    retained_samples <- metadata_aligned %>%
      dplyr::filter(.data$station_tag %in% site_order) %>%
      dplyr::group_by(.data$station_tag) %>%
      dplyr::slice_sample(n = power_resample_libraries_per_site) %>%
      dplyr::ungroup() %>%
      dplyr::arrange(
        match(.data$station_tag, site_order),
        .data$sample_code
      ) %>%
      dplyr::pull(.data$sample_code)
    tibble::tibble(
      resample_id = resample_id,
      retained_samples_list = list(retained_samples),
      retained_samples = paste(retained_samples, collapse = ";")
    )
  }
)

run_power_resample <- function(power, resample_id,
                               retained_samples_list, retained_samples) {
  current_power <- power
  reference_result <- power_network_results[[paste0("P", current_power)]]
  current_signature <- paste(
    power_cache_signature_base,
    paste0("power=", current_power),
    paste0("resample_id=", resample_id),
    paste0("samples=", retained_samples),
    sep = "||"
  )
  cache_file <- file.path(
    power_resample_cache_dir,
    paste0(
      "P", current_power, "_resample_",
      sprintf("%02d", resample_id), "_",
      digest::digest(current_signature, algo = "xxhash64"),
      ".rds"
    )
  )
  if (reuse_power_sensitivity_cache && file.exists(cache_file)) {
    cached <- readRDS(cache_file)
    if (is.list(cached) && identical(cached$signature, current_signature)) {
      message("Reusing power/resample cache P", current_power, " R", resample_id)
      return(cached$result)
    }
  }
  set.seed(power_sensitivity_seed + current_power * 100L + resample_id)
  resampled_expression <- dat_expr_g10000[
    match(retained_samples_list, rownames(dat_expr_g10000)),
    ,
    drop = FALSE
  ]
  alternative_result <- wgcna_run_validated_network(
    dat_expr = resampled_expression,
    power = current_power,
    network_type = network_type,
    correlation_type = correlation_type,
    max_p_outliers = max_p_outliers,
    minimum_module_size = minimum_module_size,
    merge_cut_height = merge_cut_height,
    deep_split = deep_split,
    pam_respects_dendro = pam_respects_dendro,
    max_block_size = max_block_size,
    verbose = 0
  )
  projected_alternative <- wgcna_orient_eigengenes(
    dat_expr_g10000,
    alternative_result$colors,
    max_p_outliers = max_p_outliers
  )$eigengenes
  result <- wgcna_match_modules(
    reference_colors = reference_result$colors,
    alternative_colors = alternative_result$colors,
    reference_eigengenes = reference_result$eigengenes,
    projected_alternative_eigengenes = projected_alternative,
    max_p_outliers = max_p_outliers
  ) %>%
    dplyr::mutate(
      power = current_power,
      resample_id = resample_id,
      retained_sample_n = length(retained_samples_list),
      .before = 1
    )
  saveRDS(
    list(signature = current_signature, result = result),
    cache_file,
    compress = FALSE
  )
  result
}

power_resampling_stability <- tidyr::expand_grid(
  power = power_sensitivity_candidates,
  resample_id = power_resample_design$resample_id
) %>%
  dplyr::left_join(power_resample_design, by = "resample_id") %>%
  purrr::pmap_dfr(run_power_resample)

power_resampling_module_summary <- power_resampling_stability %>%
  dplyr::group_by(.data$power, .data$reference_module) %>%
  dplyr::summarise(
    module_median_resampling_jaccard = stats::median(
      .data$jaccard, na.rm = TRUE
    ),
    module_minimum_resampling_jaccard = min(.data$jaccard, na.rm = TRUE),
    module_median_resampling_eigengene_agreement = stats::median(
      .data$eigengene_abs_correlation,
      na.rm = TRUE
    ),
    module_reciprocal_match_fraction = mean(
      .data$reciprocal_match, na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    candidate_moderate_resampling_module =
      .data$module_median_resampling_jaccard >=
      candidate_module_moderate_median_jaccard &
      .data$module_minimum_resampling_jaccard >=
      candidate_module_moderate_minimum_jaccard &
      .data$module_median_resampling_eigengene_agreement >=
      candidate_module_moderate_eigengene_agreement
  )

power_resampling_summary <- power_resampling_stability %>%
  dplyr::group_by(.data$power) %>%
  dplyr::summarise(
    median_resampling_jaccard = stats::median(.data$jaccard, na.rm = TRUE),
    minimum_resampling_jaccard = min(.data$jaccard, na.rm = TRUE),
    median_resampling_eigengene_agreement = stats::median(
      .data$eigengene_abs_correlation, na.rm = TRUE
    ),
    reciprocal_match_fraction = mean(.data$reciprocal_match, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::left_join(
    power_resampling_module_summary %>%
      dplyr::group_by(.data$power) %>%
      dplyr::summarise(
        candidate_moderate_resampling_module_n = sum(
          .data$candidate_moderate_resampling_module
        ),
        candidate_moderate_resampling_module_fraction = mean(
          .data$candidate_moderate_resampling_module
        ),
        .groups = "drop"
      ),
    by = "power"
  )

adjacent_power_summary_long <- dplyr::bind_rows(
  power_adjacent_preservation %>%
    dplyr::transmute(
      power = .data$reference_power,
      adjacent_pair = paste0(
        .data$reference_power, "-", .data$alternative_power
      ),
      .data$jaccard,
      .data$eigengene_abs_correlation
    ),
  power_adjacent_preservation %>%
    dplyr::transmute(
      power = .data$alternative_power,
      adjacent_pair = paste0(
        .data$reference_power, "-", .data$alternative_power
      ),
      .data$jaccard,
      .data$eigengene_abs_correlation
    )
)
adjacent_power_summary <- adjacent_power_summary_long %>%
  dplyr::group_by(.data$power) %>%
  dplyr::summarise(
    median_adjacent_power_jaccard = stats::median(
      .data$jaccard, na.rm = TRUE
    ),
    minimum_adjacent_power_jaccard = min(.data$jaccard, na.rm = TRUE),
    median_adjacent_power_eigengene_agreement = stats::median(
      .data$eigengene_abs_correlation, na.rm = TRUE
    ),
    .groups = "drop"
  )

maximum_candidate_scale_free_fit <- max(
  g10000_soft_tbl$scale_free_fit,
  na.rm = TRUE
)
final_power_selection_table <- g10000_soft_tbl %>%
  dplyr::select(
    .data$power,
    signed_scale_free_fit = .data$scale_free_fit,
    .data$slope,
    .data$mean_connectivity,
    .data$median_connectivity
  ) %>%
  dplyr::left_join(power_connectivity_distribution, by = "power") %>%
  dplyr::left_join(power_network_structure, by = "power") %>%
  dplyr::left_join(adjacent_power_summary, by = "power") %>%
  dplyr::left_join(power_resampling_summary, by = "power") %>%
  dplyr::mutate(
    scale_free_fit_deficit_from_tested_maximum =
      maximum_candidate_scale_free_fit - .data$signed_scale_free_fit,
    overly_sparse =
      .data$median_connectivity < adequate_median_connectivity_threshold,
    adequate_connectivity = !.data$overly_sparse,
    near_best_scale_free_fit =
      .data$scale_free_fit_deficit_from_tested_maximum <=
      maximum_scale_free_fit_deficit,
    adequate_resampling_stability =
      .data$candidate_moderate_resampling_module_n >= 1L,
    structurally_eligible =
      .data$adequate_connectivity &
      .data$near_best_scale_free_fit &
      .data$adequate_resampling_stability
  )

eligible_powers <- final_power_selection_table$power[
  final_power_selection_table$structurally_eligible
]
if (length(eligible_powers) == 0L) {
  stop(
    "No tested power met the predeclared connectivity, fit-deficit, and ",
    "resampling-stability criteria."
  )
}
final_selected_power <- min(eligible_powers)
final_power_selection_table <- final_power_selection_table %>%
  dplyr::mutate(
    selected = .data$power == final_selected_power,
    decision = dplyr::case_when(
      .data$selected ~ paste(
        "selected: lowest structurally eligible power after scale-free-fit",
        "improvements became marginal"
      ),
      .data$overly_sparse ~
        "not selected: median connectivity below the declared adequacy threshold",
      !.data$near_best_scale_free_fit ~
        "not selected: scale-free-fit deficit exceeds the declared tolerance",
      !.data$adequate_resampling_stability ~ paste(
        "not selected: no module met the moderate resampling membership",
        "and eigengene thresholds"
      ),
      TRUE ~ "eligible but higher than the selected power"
    ),
    selection_uses_site_contrasts = FALSE,
    selection_uses_deg_overlap = FALSE,
    selection_uses_enrichment = FALSE,
    selection_uses_chemistry = FALSE
  )

final_power_result <- power_network_results[[
  paste0("P", final_selected_power)
]]

write_csv_safe(
  final_power_selection_table,
  file.path(tables_dir, "final_power_selection_table.csv")
)
write_csv_safe(
  power_adjacent_preservation,
  file.path(tables_dir, "power_adjacent_module_preservation.csv")
)
write_csv_safe(
  power_resampling_stability,
  file.path(tables_dir, "power_site_stratified_resampling_stability.csv")
)
write_csv_safe(
  power_resampling_module_summary,
  file.path(tables_dir, "power_module_resampling_summary.csv")
)
write_csv_safe(
  power_resample_design %>% dplyr::select(-.data$retained_samples_list),
  file.path(tables_dir, "power_resampling_design.csv")
)
write_csv_safe(
  power_module_audit,
  file.path(tables_dir, "power_module_identity_audit.csv")
)
write_csv_safe(
  power_anomalous_gene_audit,
  file.path(tables_dir, "power_undersized_module_gene_audit.csv")
)
write_csv_safe(
  power_eigengene_orientation,
  file.path(tables_dir, "power_eigengene_orientation.csv")
)

knitr::kable(
  final_power_selection_table,
  digits = 4,
  caption = "Structural decision table for the 10,000-gene network power."
)
```

```{r power-decision-plot, fig.width=10, fig.height=7}
power_decision_plot <- final_power_selection_table %>%
  dplyr::select(
    .data$power,
    `Scale-free fit` = .data$signed_scale_free_fit,
    `Median connectivity (scaled)` = .data$median_connectivity,
    `Median resampling Jaccard` = .data$median_resampling_jaccard,
    `Median adjacent-power Jaccard` = .data$median_adjacent_power_jaccard
  ) %>%
  dplyr::mutate(
    `Median connectivity (scaled)` =
      `Median connectivity (scaled)` /
      max(`Median connectivity (scaled)`, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(
    cols = -"power",
    names_to = "diagnostic",
    values_to = "value"
  ) %>%
  ggplot2::ggplot(ggplot2::aes(.data$power, .data$value)) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::geom_vline(
    xintercept = final_selected_power,
    linetype = "dashed",
    colour = "#b03a2e"
  ) +
  ggplot2::facet_wrap(~diagnostic, scales = "free_y", ncol = 2) +
  ggplot2::labs(
    title = "Structural power sensitivity for the 10,000-gene network",
    subtitle = paste("Selected power:", final_selected_power),
    x = "Soft-thresholding power",
    y = "Diagnostic value"
  ) +
  theme_publication()
save_plot_bundle(
  power_decision_plot,
  file.path(plots_dir, "final_power_structural_decision"),
  width = 10,
  height = 7
)
power_decision_plot
```

# 9. Permitted evidence for network selection

```{r network-selection-evidence-policy}
network_selection_evidence_policy <- tibble::tribble(
  ~decision_domain, ~permitted_for_network_selection, ~role,
  "Connectivity and sparsity", "Yes", "primary structural criterion",
  "Module size and grey fraction", "Yes", "primary structural criterion",
  "Resampling stability", "Yes", "primary structural criterion",
  "Cross-universe preservation", "Yes", "primary structural criterion",
  "Annotation coverage", "Yes, secondary", "practical coverage criterion",
  "Number of significant site contrasts", "No", "downstream sensitivity only",
  "DEG overlap", "No", "downstream sensitivity only",
  "GO or KEGG enrichment count", "No", "downstream sensitivity only",
  "Chemistry correlation", "No", "post-selection exploratory context only"
)
write_csv_safe(
  network_selection_evidence_policy,
  file.path(tables_dir, "network_selection_evidence_policy.csv")
)
knitr::kable(
  network_selection_evidence_policy,
  caption = "Evidence hierarchy enforced for network selection."
)
```

# 10. Final-power cross-universe preservation

The cross-universe confirmation below replaces the provisional 10,000-gene
power-20 membership with the structurally selected power-14 membership. This
prevents the final gene-universe dashboard from mixing provisional and final
network statistics.

```{r final-power-cross-universe-preservation}
final_g10000_comparison_result <- list(
  network_label = paste0("G10000_P", final_selected_power),
  network_size = 10000L,
  assignment = final_power_result$assignment,
  eigengenes = tibble::as_tibble(
    final_power_result$eigengenes,
    rownames = "sample_code"
  )
)

final_universe_module_preservation <- dplyr::bind_rows(
  compare_network_pair(
    network_results$G5000,
    final_g10000_comparison_result
  ),
  compare_network_pair(
    final_g10000_comparison_result,
    network_results$G15000
  )
)

final_universe_preservation_summary <-
  final_universe_module_preservation %>%
  dplyr::group_by(
    .data$reference_network,
    .data$test_network,
    .data$reference_size,
    .data$test_size
  ) %>%
  dplyr::summarise(
    n_reference_modules = dplyr::n(),
    median_jaccard = stats::median(.data$jaccard, na.rm = TRUE),
    minimum_jaccard = min(.data$jaccard, na.rm = TRUE),
    median_reference_recovery = stats::median(
      .data$reference_recovery_fraction,
      na.rm = TRUE
    ),
    median_abs_eigengene_cor = stats::median(
      .data$eigengene_profile_abs_cor,
      na.rm = TRUE
    ),
    n_high_preservation = sum(
      .data$preservation_class == "high",
      na.rm = TRUE
    ),
    n_moderate_preservation = sum(
      .data$preservation_class == "moderate",
      na.rm = TRUE
    ),
    n_low_preservation = sum(
      .data$preservation_class == "low",
      na.rm = TRUE
    ),
    .groups = "drop"
  )

final_g10000_eligible_gene_ids <- unique(
  final_power_result$assignment$gene_id
)
final_g10000_assigned_gene_ids <- unique(
  final_power_result$assignment$gene_id[
    final_power_result$assignment$module != "grey"
  ]
)
annotated_gene_ids <- annotation_gene_tbl$gene_id[
  annotation_gene_tbl$has_readable_annotation
]
final_g10000_coverage <- tibble::tibble(
  network_label = "G10000",
  network_size = 10000L,
  eligible_gene_n = length(final_g10000_eligible_gene_ids),
  assigned_gene_n = length(final_g10000_assigned_gene_ids),
  readable_annotation_eligible_n = length(intersect(
    final_g10000_eligible_gene_ids,
    annotated_gene_ids
  )),
  readable_annotation_assigned_n = length(intersect(
    final_g10000_assigned_gene_ids,
    annotated_gene_ids
  )),
  go_mapped_eligible_n = length(intersect(
    final_g10000_eligible_gene_ids,
    go_gene_ids
  )),
  go_mapped_assigned_n = length(intersect(
    final_g10000_assigned_gene_ids,
    go_gene_ids
  ))
) %>%
  dplyr::mutate(
    readable_annotation_eligible_fraction = safe_divide(
      .data$readable_annotation_eligible_n,
      .data$eligible_gene_n
    ),
    readable_annotation_assigned_fraction = safe_divide(
      .data$readable_annotation_assigned_n,
      .data$assigned_gene_n
    ),
    go_mapped_eligible_fraction = safe_divide(
      .data$go_mapped_eligible_n,
      .data$eligible_gene_n
    ),
    go_mapped_assigned_fraction = safe_divide(
      .data$go_mapped_assigned_n,
      .data$assigned_gene_n
    )
  )

network_gene_coverage <- dplyr::bind_rows(
  network_gene_coverage %>%
    dplyr::filter(.data$network_size != 10000L),
  final_g10000_coverage
) %>%
  dplyr::arrange(.data$network_size)

final_g10000_deg_coverage <- planned_contrasts %>%
  dplyr::select(.data$contrast, .data$contrast_panel) %>%
  dplyr::mutate(
    network_label = "G10000",
    network_size = 10000L,
    total_significant_deg_n = purrr::map_int(
      .data$contrast,
      ~dplyr::n_distinct(focal_deg_tbl$gene_id[
        focal_deg_tbl$contrast == .x
      ])
    ),
    eligible_significant_deg_n = purrr::map_int(
      .data$contrast,
      ~length(intersect(
        focal_deg_tbl$gene_id[focal_deg_tbl$contrast == .x],
        final_g10000_eligible_gene_ids
      ))
    ),
    assigned_significant_deg_n = purrr::map_int(
      .data$contrast,
      ~length(intersect(
        focal_deg_tbl$gene_id[focal_deg_tbl$contrast == .x],
        final_g10000_assigned_gene_ids
      ))
    )
  ) %>%
  dplyr::mutate(
    eligible_deg_fraction = safe_divide(
      .data$eligible_significant_deg_n,
      .data$total_significant_deg_n
    ),
    assigned_deg_fraction_of_total = safe_divide(
      .data$assigned_significant_deg_n,
      .data$total_significant_deg_n
    ),
    assigned_deg_fraction_of_eligible = safe_divide(
      .data$assigned_significant_deg_n,
      .data$eligible_significant_deg_n
    )
  )

deg_coverage <- dplyr::bind_rows(
  deg_coverage %>% dplyr::filter(.data$network_size != 10000L),
  final_g10000_deg_coverage
) %>%
  dplyr::arrange(.data$network_size, .data$contrast)

write_csv_safe(
  network_gene_coverage,
  file.path(tables_dir, "annotation_and_go_coverage_by_gene_set.csv")
)
write_csv_safe(
  deg_coverage,
  file.path(tables_dir, "focal_deg_coverage_by_gene_set.csv")
)

write_csv_safe(
  final_universe_module_preservation,
  file.path(tables_dir, "final_power_module_preservation_across_gene_sets.csv")
)
write_csv_safe(
  final_universe_preservation_summary,
  file.path(tables_dir, "final_power_module_preservation_summary.csv")
)
knitr::kable(
  final_universe_preservation_summary,
  digits = 3,
  caption = "Cross-universe preservation after substituting the final 10,000-gene power network."
)
```

# 11. Site-stratified stability across gene universes

The same ten four-of-five-per-site resamples are used to compare the 5,000-,
10,000-, and 15,000-gene candidates. The 10,000-gene reference uses the
structurally selected power above; the other universes retain their independently
evaluated powers from the nested-universe analysis.

```{r gene-universe-resampling-stability, message=TRUE}
make_universe_reference <- function(network_label) {
  if (network_label == "G10000") {
    return(list(
      network_label = network_label,
      network_size = 10000L,
      power = final_selected_power,
      gene_ids = g10000_gene_ids,
      dat_expr = dat_expr_g10000,
      colors = final_power_result$colors,
      eigengenes = final_power_result$eigengenes
    ))
  }
  current_result <- network_results[[network_label]]
  current_gene_ids <- current_result$selected_gene_ids
  current_dat_expr <- t(
    expression_matrix[current_gene_ids, , drop = FALSE]
  )
  rownames(current_dat_expr) <- metadata_aligned$sample_code
  current_colors <- stats::setNames(
    as.character(current_result$assignment$module),
    as.character(current_result$assignment$gene_id)
  )[colnames(current_dat_expr)]
  me_columns_current <- grep(
    "^ME", names(current_result$eigengenes), value = TRUE
  )
  current_me <- current_result$eigengenes %>%
    dplyr::slice(match(rownames(current_dat_expr), .data$sample_code)) %>%
    dplyr::select(dplyr::all_of(me_columns_current)) %>%
    as.data.frame()
  rownames(current_me) <- rownames(current_dat_expr)
  list(
    network_label = network_label,
    network_size = current_result$network_size,
    power = as.integer(current_result$summary$selected_power[[1]]),
    gene_ids = current_gene_ids,
    dat_expr = current_dat_expr,
    colors = current_colors,
    eigengenes = current_me
  )
}

universe_reference_objects <- purrr::map(
  paste0("G", gene_set_sizes),
  make_universe_reference
)
names(universe_reference_objects) <- paste0("G", gene_set_sizes)

universe_resample_cache_dir <- file.path(
  objects_dir, "gene_universe_resample_cache"
)
dir.create(
  universe_resample_cache_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

run_universe_resample <- function(network_label, resample_id,
                                  retained_samples_list, retained_samples) {
  reference <- universe_reference_objects[[network_label]]
  if (network_label == "G10000") {
    return(
      power_resampling_stability %>%
        dplyr::filter(
          .data$power == final_selected_power,
          .data$resample_id == resample_id
        ) %>%
        dplyr::mutate(
          network_label = network_label,
          network_size = 10000L,
          .before = 1
        )
    )
  }
  current_signature <- paste(
    power_cache_signature_base,
    paste0("universe=", network_label),
    paste0("gene_hash=", digest::digest(reference$gene_ids, algo = "xxhash64")),
    paste0("power=", reference$power),
    paste0("resample_id=", resample_id),
    paste0("samples=", retained_samples),
    sep = "||"
  )
  cache_file <- file.path(
    universe_resample_cache_dir,
    paste0(
      network_label, "_resample_", sprintf("%02d", resample_id), "_",
      digest::digest(current_signature, algo = "xxhash64"), ".rds"
    )
  )
  if (reuse_power_sensitivity_cache && file.exists(cache_file)) {
    cached <- readRDS(cache_file)
    if (is.list(cached) && identical(cached$signature, current_signature)) {
      message("Reusing universe/resample cache ", network_label, " R", resample_id)
      return(cached$result)
    }
  }
  set.seed(
    power_sensitivity_seed + reference$network_size + resample_id
  )
  resampled_expression <- reference$dat_expr[
    match(retained_samples_list, rownames(reference$dat_expr)),
    ,
    drop = FALSE
  ]
  alternative <- wgcna_run_validated_network(
    dat_expr = resampled_expression,
    power = reference$power,
    network_type = network_type,
    correlation_type = correlation_type,
    max_p_outliers = max_p_outliers,
    minimum_module_size = minimum_module_size,
    merge_cut_height = merge_cut_height,
    deep_split = deep_split,
    pam_respects_dendro = pam_respects_dendro,
    max_block_size = max_block_size,
    verbose = 0
  )
  projected_alternative <- wgcna_orient_eigengenes(
    reference$dat_expr,
    alternative$colors,
    max_p_outliers = max_p_outliers
  )$eigengenes
  result <- wgcna_match_modules(
    reference_colors = reference$colors,
    alternative_colors = alternative$colors,
    reference_eigengenes = reference$eigengenes,
    projected_alternative_eigengenes = projected_alternative,
    max_p_outliers = max_p_outliers
  ) %>%
    dplyr::mutate(
      network_label = network_label,
      network_size = reference$network_size,
      power = reference$power,
      resample_id = resample_id,
      retained_sample_n = length(retained_samples_list),
      .before = 1
    )
  saveRDS(
    list(signature = current_signature, result = result),
    cache_file,
    compress = FALSE
  )
  result
}

gene_universe_resampling_stability <- tidyr::expand_grid(
  network_label = names(universe_reference_objects),
  resample_id = power_resample_design$resample_id
) %>%
  dplyr::left_join(power_resample_design, by = "resample_id") %>%
  purrr::pmap_dfr(run_universe_resample)

gene_universe_resampling_summary <- gene_universe_resampling_stability %>%
  dplyr::group_by(.data$network_label, .data$network_size, .data$power) %>%
  dplyr::summarise(
    median_resampling_jaccard = stats::median(.data$jaccard, na.rm = TRUE),
    minimum_resampling_jaccard = min(.data$jaccard, na.rm = TRUE),
    median_eigengene_agreement = stats::median(
      .data$eigengene_abs_correlation,
      na.rm = TRUE
    ),
    reciprocal_match_fraction = mean(.data$reciprocal_match, na.rm = TRUE),
    .groups = "drop"
  )

write_csv_safe(
  gene_universe_resampling_stability,
  file.path(tables_dir, "gene_universe_resampling_stability.csv")
)
write_csv_safe(
  gene_universe_resampling_summary,
  file.path(tables_dir, "gene_universe_resampling_summary.csv")
)
knitr::kable(
  gene_universe_resampling_summary,
  digits = 3,
  caption = "Site-stratified module stability across nested gene universes."
)
```

# 11. Integrated decision dashboard

The decision dashboard contains only structural criteria plus annotation
coverage as a secondary practical consideration. Site-contrast significance,
DEG recovery, and functional yield are exported in a separate downstream
sensitivity table and cannot affect the network choice.

```{r decision-dashboard}
preservation_as_reference <- final_universe_preservation_summary %>%
  dplyr::group_by(.data$reference_size) %>%
  dplyr::summarise(
    median_jaccard_to_larger = stats::median(
      .data$median_jaccard,
      na.rm = TRUE
    ),
    median_abs_eigengene_cor_to_larger = stats::median(
      .data$median_abs_eigengene_cor,
      na.rm = TRUE
    ),
    minimum_high_preservation_modules = min(
      .data$n_high_preservation,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  dplyr::rename(network_size = .data$reference_size)

contrast_dashboard <- planned_contrasts_all %>%
  dplyr::group_by(
    .data$network_label,
    .data$network_size
  ) %>%
  dplyr::summarise(
    n_fdr_supported_module_contrasts = sum(
      .data$fdr_supported,
      na.rm = TRUE
    ),
    n_large_nominal_module_contrasts = sum(
      .data$large_nominal_effect,
      na.rm = TRUE
    ),
    maximum_abs_hedges_g = max(
      abs(.data$hedges_g),
      na.rm = TRUE
    ),
    .groups = "drop"
  )

deg_dashboard <- deg_coverage %>%
  dplyr::group_by(
    .data$network_label,
    .data$network_size
  ) %>%
  dplyr::summarise(
    mean_eligible_deg_fraction = mean(
      .data$eligible_deg_fraction,
      na.rm = TRUE
    ),
    mean_assigned_deg_fraction_of_total = mean(
      .data$assigned_deg_fraction_of_total,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

decision_dashboard <- network_summary %>%
  dplyr::transmute(
    .data$network_label,
    network_size = .data$requested_gene_n,
    .data$selected_power,
    .data$selected_scale_free_fit,
    .data$selected_mean_connectivity,
    .data$selected_median_connectivity,
    .data$target_reached,
    .data$n_assigned_modules,
    .data$assigned_gene_fraction,
    .data$grey_gene_fraction,
    .data$n_blocks
  ) %>%
  dplyr::left_join(
    preservation_as_reference,
    by = "network_size"
  ) %>%
  dplyr::left_join(
    gene_universe_resampling_summary %>%
      dplyr::select(
        .data$network_label,
        .data$network_size,
        resampling_power = .data$power,
        .data$median_resampling_jaccard,
        .data$minimum_resampling_jaccard,
        .data$median_eigengene_agreement,
        .data$reciprocal_match_fraction
      ),
    by = c("network_label", "network_size")
  ) %>%
  dplyr::left_join(
    network_gene_coverage,
    by = c("network_label", "network_size")
  ) %>%
  dplyr::mutate(
    selected_power = dplyr::if_else(
      .data$network_size == 10000L,
      as.numeric(final_power_selection_table$power[
        final_power_selection_table$selected
      ]),
      as.numeric(.data$selected_power)
    ),
    selected_scale_free_fit = dplyr::if_else(
      .data$network_size == 10000L,
      final_power_selection_table$signed_scale_free_fit[
        final_power_selection_table$selected
      ],
      .data$selected_scale_free_fit
    ),
    selected_mean_connectivity = dplyr::if_else(
      .data$network_size == 10000L,
      final_power_selection_table$mean_connectivity[
        final_power_selection_table$selected
      ],
      .data$selected_mean_connectivity
    ),
    selected_median_connectivity = dplyr::if_else(
      .data$network_size == 10000L,
      final_power_selection_table$median_connectivity[
        final_power_selection_table$selected
      ],
      .data$selected_median_connectivity
    ),
    target_reached = dplyr::if_else(
      .data$network_size == 10000L,
      final_power_selection_table$signed_scale_free_fit[
        final_power_selection_table$selected
      ] >= soft_threshold_target,
      .data$target_reached
    ),
    n_assigned_modules = dplyr::if_else(
      .data$network_size == 10000L,
      as.integer(final_power_selection_table$module_n[
        final_power_selection_table$selected
      ]),
      as.integer(.data$n_assigned_modules)
    ),
    assigned_gene_fraction = dplyr::if_else(
      .data$network_size == 10000L,
      final_power_selection_table$assigned_gene_fraction[
        final_power_selection_table$selected
      ],
      .data$assigned_gene_fraction
    ),
    grey_gene_fraction = dplyr::if_else(
      .data$network_size == 10000L,
      final_power_selection_table$grey_fraction[
        final_power_selection_table$selected
      ],
      .data$grey_gene_fraction
    ),
    n_blocks = dplyr::if_else(
      .data$network_size == 10000L,
      as.integer(final_power_selection_table$n_blocks[
        final_power_selection_table$selected
      ]),
      as.integer(.data$n_blocks)
    )
  ) %>%
  dplyr::arrange(.data$network_size)

downstream_sensitivity_dashboard <- contrast_dashboard %>%
  dplyr::left_join(
    deg_dashboard,
    by = c("network_label", "network_size")
  ) %>%
  dplyr::mutate(
    permitted_for_network_selection = FALSE,
    role = "downstream sensitivity only"
  ) %>%
  dplyr::arrange(.data$network_size)

write_csv_safe(
  decision_dashboard,
  file.path(tables_dir, "wgcna_gene_set_decision_dashboard.csv")
)
write_csv_safe(
  downstream_sensitivity_dashboard,
  file.path(tables_dir, "wgcna_downstream_sensitivity_dashboard.csv")
)

knitr::kable(
  decision_dashboard,
  digits = 4,
  caption = "Structural diagnostics for selecting the final WGCNA gene set."
)
knitr::kable(
  downstream_sensitivity_dashboard,
  digits = 4,
  caption = "Downstream results excluded from network selection."
)
```

## Selected network for the downstream exploratory rerun

The 10,000-gene universe is selected as the smallest expanded network that
retains much of the 5,000-gene structure without the greater fragmentation and
weaker incremental preservation of the 15,000-gene expansion. Its improved
annotation coverage is a secondary practical benefit. DEG recovery, site
significance, enrichment yield, and chemistry do not enter this decision.

```{r selected-network-handoff}
selected_gene_set_size <- 10000L
selected_network_label <- paste0("G", selected_gene_set_size)

if (!selected_network_label %in% names(network_results)) {
  stop("The selected WGCNA gene-set size was not reconstructed.")
}

selected_network_result <- final_power_result
selected_network_summary <- final_power_selection_table %>%
  dplyr::filter(.data$selected) %>%
  dplyr::transmute(
    selected_power = .data$power,
    selected_power_rule = .data$decision,
    selected_scale_free_fit = .data$signed_scale_free_fit,
    selected_slope = .data$slope,
    selected_mean_connectivity = .data$mean_connectivity,
    selected_median_connectivity = .data$median_connectivity,
    target_reached = .data$signed_scale_free_fit >= soft_threshold_target,
    n_blocks = .env$final_power_result$n_blocks
  )

selected_module_membership <- selected_network_result$assignment %>%
  dplyr::mutate(
    network_label = selected_network_label,
    network_size = selected_gene_set_size,
    .before = 1
  )

selected_kme_tbl <- tibble::as_tibble(
  WGCNA::signedKME(
    dat_expr_g10000,
    selected_network_result$eigengenes,
    outputColumnName = "kME",
    corFnc = correlation_type,
    corOptions = paste0("maxPOutliers = ", max_p_outliers)
  ),
  rownames = "gene_id"
)
selected_module_membership$module_kME_signed <- NA_real_
for (current_module in unique(selected_module_membership$module)) {
  current_column <- paste0("kME", current_module)
  current_rows <- selected_module_membership$module == current_module
  if (current_column %in% names(selected_kme_tbl)) {
    selected_module_membership$module_kME_signed[current_rows] <-
      selected_kme_tbl[[current_column]][match(
        selected_module_membership$gene_id[current_rows],
        selected_kme_tbl$gene_id
      )]
  }
}
selected_module_membership <- selected_module_membership %>%
  dplyr::mutate(
    module_kME_abs = abs(.data$module_kME_signed)
  )

selected_module_summary <- selected_network_result$module_summary %>%
  dplyr::select(
    .data$module,
    .data$n_genes,
    .data$module_type,
    .data$module_label
  )
selected_module_eigengenes_by_sample <- tibble::as_tibble(
  selected_network_result$eigengenes,
  rownames = "sample_code"
) %>%
  dplyr::left_join(
    metadata_aligned %>%
      dplyr::select(
        .data$sample_code,
        .data$site_id,
        .data$station_tag,
        dplyr::any_of(c("display_label", "site_name", "station_order"))
      ),
    by = "sample_code"
  )
selected_me_columns <- grep(
  "^ME",
  names(selected_module_eigengenes_by_sample),
  value = TRUE
)
selected_site_columns <- intersect(
  c(
    "site_id",
    "station_tag",
    "display_label",
    "site_name",
    "station_order"
  ),
  names(selected_module_eigengenes_by_sample)
)
selected_module_eigengenes_by_site <-
  selected_module_eigengenes_by_sample %>%
  dplyr::group_by(dplyr::across(dplyr::all_of(selected_site_columns))) %>%
  dplyr::summarise(
    dplyr::across(
      dplyr::all_of(selected_me_columns),
      ~mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  )

selected_soft_threshold_diagnostics <-
  network_results$G10000$soft
selected_soft_threshold_decision <- tibble::tibble(
  selected_power = selected_network_summary$selected_power,
  rule = selected_network_summary$selected_power_rule,
  target = soft_threshold_target,
  network_type = network_type,
  target_reached = selected_network_summary$target_reached
)

selected_membership_file <- file.path(
  tables_dir,
  "selected_wgcna_gene_module_membership.csv"
)
selected_eigengene_sample_file <- file.path(
  tables_dir,
  "selected_wgcna_module_eigengenes_by_sample.csv"
)
selected_eigengene_site_file <- file.path(
  tables_dir,
  "selected_wgcna_module_eigengenes_by_site.csv"
)
selected_module_summary_file <- file.path(
  tables_dir,
  "selected_wgcna_module_summary.csv"
)
selected_soft_diagnostics_file <- file.path(
  tables_dir,
  "selected_wgcna_soft_threshold_diagnostics.csv"
)
selected_soft_decision_file <- file.path(
  tables_dir,
  "selected_wgcna_soft_threshold_decision.csv"
)
selected_orientation_file <- file.path(
  tables_dir,
  "selected_wgcna_eigengene_orientation.csv"
)
selected_module_identity_audit_file <- file.path(
  tables_dir,
  "selected_wgcna_module_identity_audit.csv"
)

write_csv_safe(selected_module_membership, selected_membership_file)
write_csv_safe(
  selected_module_eigengenes_by_sample,
  selected_eigengene_sample_file
)
write_csv_safe(
  selected_module_eigengenes_by_site,
  selected_eigengene_site_file
)
write_csv_safe(selected_module_summary, selected_module_summary_file)
write_csv_safe(
  selected_soft_threshold_diagnostics,
  selected_soft_diagnostics_file
)
write_csv_safe(
  selected_soft_threshold_decision,
  selected_soft_decision_file
)
write_csv_safe(
  selected_network_result$orientation,
  selected_orientation_file
)
write_csv_safe(
  selected_network_result$module_audit,
  selected_module_identity_audit_file
)

valid_selected_sizes <- selected_module_summary$n_genes[
  selected_module_summary$module != "grey"
]
if (nrow(selected_module_membership) != selected_gene_set_size) {
  stop("Final requested gene count does not equal retained network gene count.")
}
if (any(valid_selected_sizes < minimum_module_size)) {
  stop("The selected network contains an undersized assigned module.")
}
if (!setequal(
  paste0("ME", sort(unique(selected_module_membership$module))),
  selected_me_columns
)) {
  stop("Selected module-colour labels do not match eigengene labels.")
}

selection_evidence <- tibble::tibble(
  criterion = c(
    "G5000_to_G10000_median_Jaccard",
    "G5000_to_G10000_median_abs_eigengene_correlation",
    "G10000_to_G15000_median_Jaccard",
    "G10000_to_G15000_low_preservation_modules",
    "G5000_median_resampling_Jaccard",
    "G10000_median_resampling_Jaccard",
    "G15000_median_resampling_Jaccard",
    "G5000_GO_mapped_eligible_genes",
    "G10000_GO_mapped_eligible_genes",
    "G15000_GO_mapped_eligible_genes",
    "G10000_selected_power",
    "G10000_selected_power_median_connectivity",
    "G10000_selected_power_median_resampling_Jaccard"
  ),
  value = c(
    final_universe_preservation_summary$median_jaccard[
      final_universe_preservation_summary$reference_size == 5000L &
        final_universe_preservation_summary$test_size == 10000L
    ],
    final_universe_preservation_summary$median_abs_eigengene_cor[
      final_universe_preservation_summary$reference_size == 5000L &
        final_universe_preservation_summary$test_size == 10000L
    ],
    final_universe_preservation_summary$median_jaccard[
      final_universe_preservation_summary$reference_size == 10000L &
        final_universe_preservation_summary$test_size == 15000L
    ],
    final_universe_preservation_summary$n_low_preservation[
      final_universe_preservation_summary$reference_size == 10000L &
        final_universe_preservation_summary$test_size == 15000L
    ],
    gene_universe_resampling_summary$median_resampling_jaccard[
      gene_universe_resampling_summary$network_size == 5000L
    ],
    gene_universe_resampling_summary$median_resampling_jaccard[
      gene_universe_resampling_summary$network_size == 10000L
    ],
    gene_universe_resampling_summary$median_resampling_jaccard[
      gene_universe_resampling_summary$network_size == 15000L
    ],
    network_gene_coverage$go_mapped_eligible_n[
      network_gene_coverage$network_size == 5000L
    ],
    network_gene_coverage$go_mapped_eligible_n[
      network_gene_coverage$network_size == 10000L
    ],
    network_gene_coverage$go_mapped_eligible_n[
      network_gene_coverage$network_size == 15000L
    ],
    final_selected_power,
    final_power_selection_table$median_connectivity[
      final_power_selection_table$selected
    ],
    final_power_selection_table$median_resampling_jaccard[
      final_power_selection_table$selected
    ]
  ),
  permitted_for_selection = c(
    rep(TRUE, 7),
    rep("secondary", 3),
    rep(TRUE, 3)
  )
)
write_csv_safe(
  selection_evidence,
  file.path(tables_dir, "selected_wgcna_network_evidence.csv")
)

project_relative_path <- function(path) {
  absolute_path <- normalizePath(path, winslash = "/", mustWork = TRUE)
  project_prefix <- paste0(project_dir, "/")
  if (startsWith(absolute_path, project_prefix)) {
    substring(absolute_path, nchar(project_prefix) + 1L)
  } else {
    absolute_path
  }
}

selected_network_manifest <- tibble::tibble(
  selection_status = "final_selected_for_step08",
  selection_rule = paste(
    "10,000 genes selected from structural cross-universe preservation,",
    "resampling stability, connectivity and module assignment; annotation",
    "coverage used only secondarily; site, DEG, enrichment and chemistry",
    "results excluded. Power is the lowest structurally eligible candidate"
  ),
  selected_network_label = selected_network_label,
  selected_gene_set_size = selected_gene_set_size,
  selected_power = selected_network_summary$selected_power,
  soft_threshold_target = soft_threshold_target,
  soft_threshold_target_reached = selected_network_summary$target_reached,
  network_type = network_type,
  correlation_type = correlation_type,
  max_p_outliers = max_p_outliers,
  minimum_module_size = minimum_module_size,
  merge_cut_height = merge_cut_height,
  deep_split = deep_split,
  pam_respects_dendro = pam_respects_dendro,
  max_block_size = max_block_size,
  n_blocks = selected_network_summary$n_blocks,
  wgcna_version = as.character(utils::packageVersion("WGCNA")),
  eigengene_orientation_rule =
    "positive correlation with average standardized member-gene expression",
  undersized_module_policy =
    "reclassify post-blockwise non-grey groups below minModuleSize to exact grey",
  gene_membership_file = project_relative_path(selected_membership_file),
  module_eigengenes_by_sample_file = project_relative_path(
    selected_eigengene_sample_file
  ),
  module_eigengenes_by_site_file = project_relative_path(
    selected_eigengene_site_file
  ),
  module_summary_file = project_relative_path(selected_module_summary_file),
  soft_threshold_diagnostics_file = project_relative_path(
    selected_soft_diagnostics_file
  ),
  soft_threshold_decision_file = project_relative_path(
    selected_soft_decision_file
  ),
  eigengene_orientation_file = project_relative_path(selected_orientation_file),
  module_identity_audit_file = project_relative_path(
    selected_module_identity_audit_file
  ),
  final_power_selection_file = project_relative_path(
    file.path(tables_dir, "final_power_selection_table.csv")
  )
)
write_csv_safe(
  selected_network_manifest,
  file.path(tables_dir, "selected_wgcna_network_manifest.csv")
)

knitr::kable(
  selected_network_manifest %>%
    dplyr::select(
      .data$selected_network_label,
      .data$selected_gene_set_size,
      .data$selected_power,
      .data$soft_threshold_target_reached,
      .data$n_blocks,
      .data$max_block_size
    ),
  caption = "WGCNA network selected for the final site-associated module analysis."
)
knitr::kable(
  selection_evidence,
  digits = 4,
  caption = "Evidence retained with the selected-network decision."
)
```

```{r decision-notes, results='asis'}
cat("## Reading the comparison\n\n")

if (all(!network_summary$target_reached)) {
  cat(
    "None of the tested gene sets reached the prespecified scale-free-topology target. ",
    "The final network should therefore not be described as scale-free, and gene-set selection should rely on the combined diagnostics rather than on this statistic alone.\n\n",
    sep = ""
  )
} else {
  reached_sizes <- network_summary$requested_gene_n[
    network_summary$target_reached
  ]
  cat(
    "The scale-free-topology target was reached by: ",
    paste(format(reached_sizes, big.mark = ","), collapse = ", "),
    " genes. This remains one network diagnostic rather than a sufficient selection criterion.\n\n",
    sep = ""
  )
}

if (nrow(final_universe_preservation_summary) > 0L) {
  best_preservation <- final_universe_preservation_summary %>%
    dplyr::arrange(
      dplyr::desc(.data$median_jaccard),
      dplyr::desc(.data$median_abs_eigengene_cor)
    ) %>%
    dplyr::slice_head(n = 1)

  cat(
    "The strongest overall cross-size module preservation occurred for the ",
    format(best_preservation$reference_size, big.mark = ","),
    " to ",
    format(best_preservation$test_size, big.mark = ","),
    " comparison (median Jaccard = ",
    sprintf("%.3f", best_preservation$median_jaccard),
    "; median absolute eigengene correlation = ",
    sprintf("%.3f", best_preservation$median_abs_eigengene_cor),
    ").\n\n",
    sep = ""
  )
}

cat(
  "The 10,000-gene universe is selected from structural evidence: cross-universe preservation, site-stratified resampling stability, connectivity, assignment, and module-size behaviour. Annotation coverage is secondary. Site-contrast significance, DEG recovery, enrichment yield, and chemistry were excluded from selection. Within the 10,000-gene universe, power ",
  final_selected_power,
  " is the lowest candidate meeting the declared connectivity, scale-free-fit-deficit, and resampling-stability criteria. Step 08 consumes this explicit final-network handoff.\n",
  sep = ""
)
```

# 9. Save the lightweight comparison object

```{r save-lightweight-object}
lightweight_results <- list(
  settings = list(
    gene_set_sizes = gene_set_sizes,
    candidate_powers = candidate_powers,
    soft_threshold_target = soft_threshold_target,
    reference_power_for_comparison = reference_power_for_comparison,
    network_type = network_type,
    correlation_type = correlation_type,
    max_p_outliers = max_p_outliers,
    minimum_module_size = minimum_module_size,
    merge_cut_height = merge_cut_height,
    deep_split = deep_split,
    pam_respects_dendro = pam_respects_dendro,
    max_block_size = max_block_size,
    planned_contrasts = planned_contrasts
  ),
  network_summary = network_summary,
  soft_threshold = soft_threshold_all,
  module_summary = module_summary_all,
  module_assignments = purrr::map(
    network_results,
    "assignment"
  ),
  eigengenes = purrr::map(
    network_results,
    "eigengenes"
  ),
  planned_contrasts = planned_contrasts_all,
  module_preservation = module_preservation,
  contrast_preservation = contrast_preservation,
  annotation_coverage = network_gene_coverage,
  deg_coverage = deg_coverage,
  decision_dashboard = decision_dashboard
  , power_selection = final_power_selection_table
  , power_adjacent_preservation = power_adjacent_preservation
  , power_resampling_stability = power_resampling_stability
  , gene_universe_resampling_stability = gene_universe_resampling_stability
  , final_power_module_preservation = final_universe_module_preservation
  , final_power_preservation_summary = final_universe_preservation_summary
)

saveRDS(
  lightweight_results,
  file.path(
    objects_dir,
    "wgcna_gene_set_sensitivity_lightweight.rds"
  ),
  compress = "xz"
)
```

# 10. Output inventory and session information

```{r output-inventory}
output_inventory <- tibble::tibble(
  file = list.files(
    results_dir,
    recursive = TRUE,
    full.names = TRUE
  )
) %>%
  dplyr::mutate(
    relative_path = stringr::str_remove(
      .data$file,
      paste0("^", stringr::fixed(results_dir), "/?")
    ),
    size_bytes = file.info(.data$file)$size
  ) %>%
  dplyr::select(
    .data$relative_path,
    .data$size_bytes
  ) %>%
  dplyr::arrange(.data$relative_path)

write_csv_safe(
  output_inventory,
  file.path(tables_dir, "output_inventory.csv")
)

knitr::kable(
  output_inventory,
  caption = "Outputs generated by the WGCNA gene-set sensitivity report."
)
```

```{r session-info}
session_info_path <- file.path(logs_dir, "session_info.txt")
writeLines(capture.output(sessionInfo()), session_info_path)
sessionInfo()
```

